### 2025-04-21 rebuttal.tex
Started convex relaxation based solver (MA thesis) paper

\documentclass[10pt,twocolumn,letterpaper]{article}
\usepackage[rebuttal]{cvpr}

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}

% Import additional packages in the preamble file, before hyperref
\input{preamble}

% If you comment hyperref and then uncomment it, you should delete
% egpaper.aux before re-running latex.  (Or just hit 'q' on the first latex
% run, let it finish, and you should be clear).
\definecolor{cvprblue}{rgb}{0.21,0.49,0.74}
\usepackage[pagebackref,breaklinks,colorlinks,allcolors=cvprblue]{hyperref}

% If you wish to avoid re-using figure, table, and equation numbers from
% the main paper, please uncomment the following and change the numbers
% appropriately.
%\setcounter{figure}{2}
%\setcounter{table}{1}
%\setcounter{equation}{2}

% If you wish to avoid re-using reference numbers from the main paper,
% please uncomment the following and change the counter value to the
% number of references you have in the main paper (here, 100).
%\makeatletter
%\apptocmd{\thebibliography}{\global\c@NAT@ctr 100\relax}{}{}
%\makeatother

%%%%%%%%% PAPER ID  - PLEASE UPDATE
\def\paperID{16006} % *** Enter the Paper ID here
\def\confName{CVPR}
\def\confYear{2025}

\input{sec/custom_commands}
\begin{document}

%%%%%%%%% TITLE - PLEASE UPDATE
\title{Towards certifiable outlier-robust point cloud registration in real-time using intersection graphs}  % **** Enter the paper title here

\maketitle
\thispagestyle{empty}
\appendix

We first want to thank the reviewers for their comments, feedback and valuable ideas.
We are happy to hear that reviewer 1 enjoyed reading our paper, and that reviewer 1 and 3 think the ideas in our paper may inspire the research community.

\textbf{Response to reviewer 1}. To address the overall confusion about the problem setting reviewer 1 seemingly had, we realize that we will need to elaborate more on our specific use case which is global localization for autonomous driving, using LiDAR-sensors. For this use-case, only rigid transformation is needed (we will change the title to mention rigid), and in combination with local feature descriptors, a few hundred points are usually enough. We agree here fully with reviewer 1 that different use-cases have different requirements on scalability and speed. We think that considering there is a high-frequency odometry, doing LiDAR-registration with 10Hz (100ms) is the bare-minimum that could still count as real-time in our use-case.
The assumption of pre-made correspondences indeed means that we cannot just plug in raw point clouds and is therefore a limitation. Instead, we need to use keypoints and match them with feature descriptors, this was done in the evaluation section.
We fully agree with reviewer 1 that knowing what influence epsilon has on the estimation error \cite{antonante2024-robust-perception} would be valuable, we will add an evaluation for this.

\textbf{Response to reviewer 2}.
Addressing the main concern of novelty of reviewer 2, we want to point out that while on the surface, our approach looks similar to \cite{zhang20233d} because it uses maximal cliques, our approach is mathematically rigorous by stating it as an optimization problem with an exactly defined objective function, namely the Truncated Least Squares (TLS). It offers in our opinion a great insight why listing maximal cliques works well in practice, instead of only empirically observing this fact. Neither was it previously described what compatibility criterion really means (namely pairwise intersection of the so-called T-sets),
nor what the connection is between the global minimas of the Truncated Least Squares (TLS) objective and the maximal cliques of the newly introduced intersection graph. Also, when talking about global optimality, we have shown that maximum cliques are not a sufficient condition for global optimality due to collections of SE3- T-sets lacking the Helly-property, a key insight about the combinatorics of the problem that was previously overlooked.
\textbf{Question 2}: The answer is unfortunately no, as we wrote at the end of section 5.1.2.\\
%\textit{Unfortunately, being only able to reject some and not all instances means our solver is currently neither globally optimal (being able to find the global minimum of every instance and outputting a valid lower bound on every instance), nor certifiable (outputting only a valid lower bound on every instance)}.\\
\textbf{Question 3}: GNC definitively undermines global optimality because it is a heuristic, yes. But we really mean GNC often \textit{finds} the global optimum, not that it \textit{proves} that the found solution is the global optimum.\\
\textbf{Question 4}: We indeed have a bad performance but only with the FPFH descriptor on 3DMatch, we will investigate the root cause of this. Since we have a good performance on 3DMatch using FCGF descriptor (83.92\%) and on KITTI even an accuracy comparable to state-of-the art with both FPFH and FCGF (95.5\% and 96.76\% respectively, top-three overall), this shows our algorithm is able to compete with state-of-the-art with both descriptors at least on KITTI. We suspect a wrong parameter in the config or a bug in one of the evaluation scripts for which we ran out of time for fixing. \textbf{Question 5.1}: We will report the runtime on the real datasets KITTI and 3DMatch, we indeed forgot to include them. \textbf{Question 5.2}: We will consider adding synthetic experiments of common other registration algorithms. The purpose of the evaluation section 5.1 is mainly for comparing the only available globally optimal solver STRIDE regarding registration accuracy. Also, the solver STRIDE is extremely slow and not scalable at all for more than 100 points, this is why we cannot use it in the experiments 5.2 and 5.3.
\textbf{Question 5.3}: We will attempt to do at least 10, maybe 50.\\
\textbf{Question 5.4}: Due to random subsampling, using less points may only make our algorithm look less accurate compared to the others, not more accurate. But we fully agree that the experiments should be done under the same conditions for comparability. We will attempt to reproduce all the numbers ourselves under the same conditions instead of reporting from \cite{zhang20233d}.
We will add qualitative results (i.e. example of registered point clouds) if space permits.

\textbf{Response to reviewer 3}. We indeed did not tune the parameters and still got top-three results on KITTI compared to SoTa. We are confident that we will find the issue with the FPFH descriptor on 3DMatch, and obtain overall more consistent evaluation results that we will also discuss more thoroughly. We will add the info how much time each submodule of the algorithm takes.
Regarding Figure 1 missing explanations, we indeed see now that identifiers like $\vc_i$, $\vt$ and $\epsilon$ are not present in the illustration, we apologize for the confusion. We do not see any problem with neither Eq. (10) nor Eq. (11), could you please elaborate ?

%%%%%%%%% REFERENCES
{
    \small
    \bibliographystyle{ieeenat_fullname}
    \bibliography{main}
}

\end{document}

### 2025-04-21 sec/0_abstract.tex
Started convex relaxation based solver (MA thesis) paper

\begin{abstract}
Outlier-robust point cloud registration where correspondences are given but up to 90\% of them may be outliers is commonly formulated as as the \textit{Maximum Consensus} (MC) or \textit{Truncated Least Squares} (TLS) optimization problem.
This formulation has been shown to be very robust to outliers but actually solving the underlying combinatorial optimization problem exactly, i.e. with global optimality guarantee, remains challenging. Approaches using semidefinite relaxations are not well scalable albeit running in polynomial time and instead have a runtime on the order of hundreds of seconds for 100 points. O

We propose a new convex relaxation that runs in linear time and can be used in combination with Branch-and-Bound. We present results on synchetic data for the rotation-only (Wahba) problem and translation problem separately.

\end{abstract}

### 2025-04-21 sec/1_my_intro.tex
Started convex relaxation based solver (MA thesis) paper

\section{Introduction}
\label{sec:intro}

Outlier-robust point cloud registration is a fundamental and long-studied problem, it is the core of many applications such as scene reconstruction or localization in autonomous driving. Most available methods solve the problem satisfactorily but provide no guarantees -- they are hard instances of the problem, either with many outliers (e.g. 90\%) or symmetries in the scene. On such instances, point cloud registration often fails, outputting an arbitrarily wrong solution.

From a statistical point of view, this is essentially a robust regression problem, more specifically a best-subset selection. This means, we need to find the best subset of data points as well as the model parameters that best fit this subset.
The model is in this case a rigid 3D-transformation, the model parameter a rigid transform $\in \SEthree$. Each data point is a pair of corresponding 3D points of from both point clouds.

In summary, our main contributions are (TODO): 
\begin{enumerate}
	\item A convex relaxation for the Truncated Least Squares objective function that can be computed in linear time
	\item Computationally very efficient custom solvers for minimizing the convex relaxation over translations and rotations
\end{enumerate}

### 2025-04-21 sec/2_related_work.tex
Started convex relaxation based solver (MA thesis) paper

\section{Related Work}
\label{sec:related-work}

Globally optimal and certifiable algorithms have been studied in computer vision \cite{10.1007/978-3-030-58539-6_18, 10023976, Convex-Relaxations-for-Pose-Graph-Optimization-With-Outliers} and robotics \cite{7759681}, both for point cloud registration (both robust and outlier-free \cite{8100078, GARCIASALGUERO2023103862}) and other problems in computer vision \cite{10378410}, but remained impractical for real-time applications because they were too slow. In the following, we briefly review algorithms for solving the robust point cloud registration problem with known correspondences, using either Truncated Least Squares (TLS) or the closely related Maximum Consensus (MC) formulation. We review only traditional methods even though deep learning approaches have been recently successful as well (e.g. \cite{9157005, bai2021pointdsc}), since neural networks greatly add to the difficulty of giving guarantees about the solution.

\subsection{Heuristic methods}
Algorithms based on randomly sampling a subset of correspondences and checking whether they are all inliers, i.e. RANSAC were used for a long time. Torr et al. \cite{TORR2000138} proposed with MLESAC not only to maximize the number of inliers, but also to consider the residuals, an idea similar to the Truncated Least Squares objective. Recently, variants of RANSAC that consider geometric compatibility have been proposed \cite{9052691, 9552513}, improving outlier robustness up to 95\%.
Yang et al. \cite{Yang2020OneRT} presented TEASER++, a heuristic algorithm based on the TLS objective that provides many algorithmic insights and popularized the idea of finding cliques in compatibility graphs. They first construct a graph of compatible measurements and then prune outliers by finding the maximum clique. They then apply a dimensionality decomposition idea where they first solve for rotation using \textit{Graduated Non-Convexity} (GNC) and then separately over different dimensions of the translation using \textit{adaptive voting}. Separately from this heuristic algorithm, they propose a second globally optimal algorithm that uses semidefinite relaxations. However, it can only prove global optimality (i.e. \textit{certify}) for one subproblem at a time, i.e. rotation or translation, not the full 3D-transform. Finally, their algorithm is not globally optimal, since their proof requires the assumption that an instance is not "too hard", i.e. they make a priori unverifiable assumptions about the instance [Assump. 2, 3 in Thm.17]\cite{Yang20tro-teaser}.

Methods using compatibility graphs are popular for point cloud registration \cite{Yang20tro-teaser, SC2-PCR-Chen-2022-CVPR, 10161215, lim2024kissmatcherfastrobustpoint} and outlier removal \cite{7410607, 10091912, zhang20233d}. They either list all maximal cliques, or search for the largest maximal clique (the \textit{maximum} clique), which is generally NP-Hard \cite[p.15]{approx-algo-book-shmoys-et-al}. Some approaches such as ROBIN \cite{9562007} or CLIPPER \cite{10432947} approximate the maximum clique.
Recently, the concept of dimensionality decomposition has gained popularity. The idea is to decompose the 6DoF problem into rotation and translation \cite{Yang2020OneRT, 9485090}, or even more \cite{9878458, 10656079} subproblems, which are then solved sequentially.
However, solving these subproblems sequentially (even to global optimality) has not been shown to be equivalent to solving the original optimization problem -- as recently hinted, it very likely breaks global optimality \cite{9878458}.

Other optimization methods for minimizing the TLS objective have been proposed, but they are either only heuristic, such as \cite{Barratt2020}, or not practical for higher dimensions \cite{doi:10.1080/10618600.2017.1390471}.

\subsection{Exact methods}

Exact methods that guarantee to find the global optimum are mostly based on Branch-and-Bound (BnB) or semidefinite relaxations. A common formulation, closely related to the TLS objective, is the
Maximum Consensus (MC), one where the number of inliers is maximized. Branch-and-Bound has been proposed for rotation-only solving of this problem \cite{4408896, 10.1007/978-3-642-37444-9_42}. Globally optimal methods based on BnB have usually been slow, even when searching over a 2D rotation \cite{9447984} using nested BnB approaches.

Another class of methods uses tight convex relaxations based on Semidefinite Programs (SDPs). In their follow-up work, Yang et al. \cite{9785843} propose a sparse SPD relaxation for (among other problems) the full 6-DoF TLS point cloud registration problem and also a fast rank-1 SDP solver \textit{STRIDE}. They can solve synthetic instances with $N=100$ points to global optimality with up to 80\% outlier rate, but require over 200 seconds. SDPs have been used for similar problems such as multiple point cloud registration \cite{9157383}, but with similarly large runtimes.

### 2025-04-21 sec/3_tls_opt.tex
Started convex relaxation based solver (MA thesis) paper

\section{Problem formulation}
The problem is to compute a 3D-transform $(\MR^*, \vt^*) \in \SEthree$ that optimally aligns two point sequences $\mathcal{P} = \left(\vp_1, ..., \vp_N\right), \mathcal{Q} = \left(\vq_1, ..., \vq_N\right) \in \RthreeByN$ of noisy measurements. We know the correspondences between points ($\vp_i$ and $\vq_i$), but a large fraction (up to 95\%) can be outliers, as it is typical for feature-based matching methods.
\subsection{The Truncated Least Squares problem}
The Truncated Least Squares (TLS) problem is a least squares esitmation problem where residuals are truncadet above a threshold. (It is similar to the Maximum Consensus problem but additionally considers the residual values, obtaining slightly better estimates.)
Outliers are handled by truncating residuals above a threshold $\epsilon^2$, yielding the \textit{Truncated Least Squares} (TLS) problem:

\begin{figure}[!ht]
	\centering
	\begin{adjustbox}{width=1.\linewidth}
		\input{figures/tls_cost_many_term.pgf}
	\end{adjustbox}
	\caption{The non-convex Truncated least-squares (TLS) Funktion with multiple local minima.Shown is the function  $\sum_{i=1}^{N}\min((x - y_i)^2, \epsilon^2)$ for $\epsilon^2=0.1$, and nine $y_i$'s (gray lines).}
	\label{fig:tlscostmulterm}
\end{figure}

Die truncated least squares Funktion ist eine Funktion der Form:

\begin{equation}
	\begin{aligned}
		\text{TLS}(x) = \sum_{i=1}^{N}\min(r(x, y_i)^2, \epsilon^2)
	\end{aligned}
\end{equation}

where $r(\cdot)$ is a residual function like $\norm{x - \vy_i}$ measuring the deviation of the model given the parameters $x$ and the data points $\vy_i$. This function is shown in Fig. \ref{fig:tlscostmulterm}.

The point cloud registration problem in the TLS formulation is therefore:

\begin{equation}
	\label{eq:pcr-tls}
	\begin{aligned}
		(\MR^*, \vt^*) =  \argmin_{(\MR, \vt) \in \SEthree}  \sum_{i=1}^{N} \min \left(\normsq{\MR \vp_i  - \vq_i  + \vt}, \epsilon^2 \right)
	\end{aligned}
\end{equation}

Among the many ways to make least squares regression robust to outliers \cite[Ch. 3]{elements-of-stats-learning-book}, the TLS formulation has recently gained populariy \cite{Yang20tro-teaser, 9785843, NIPS2017_9f53d83e, doi:10.1080/10618600.2017.1390471, NIPS2010_01882513} since it has been shown to be very robust even to high outlier rates of 90\% \cite{Yang20tro-teaser}.
This problem is a non-convex combinatorial problem that is difficult to solve (requiring non-convex binary constraints \cite{5459398}).

The parameter $\epsilon$ is best understood as the known measurement accuracy of each point. It can be estimated by collecting samples, i.e. in a data-driven manner \cite[p.8]{Chin2017TheMC}.

\section{Wheighted Least Squares convex relaxation}

In this section we propose a novel convex relaxation for the Truncated Least Squares problem that can be computed in linear time.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.6\linewidth]{figures/wls_relax_idea_with_relax}
	\caption{The idea of the WLS convex relaxation: We compute the maximum $r_{max}$ of each convex term (orange) over the given interval (blue) and then reweight it such that it remains below $\epsilon^2$ over this interval and is thus not truncated over the entire interval (green).}
	\label{fig:wlsrelaxideawithrelax}
\end{figure}

First, consider the truncated least squares objective for a general residual term $r(\cdot)$ that depends on the model (i.e rotation or translation):

\begin{equation}
	\label{eq:tls-scalar-for-wls}
	\begin{aligned}
		\text{TLS}(\mathbf{x}) = \sum_{i=1}^{N} \min \left(r(\mathbf{x}, \mathbf{x}_i)^2, \epsilon^2 \right)\\
	\end{aligned}
\end{equation}

where $\mathbf{x}_i$ are data points and $r(\cdot)^2$ is some convex residual function such as the squared Euclidean distance $||\mathbf{x} - \mathbf{x}_i||_2^2$. In the following, we will refer to the i-th residual term $r(\mathbf{x}, \mathbf{x_i})^2$ simply as $r_i$.
The idea of the WLS-relaxation is the observation that this function is only non-convex because of the truncation -- the residuals are convex (parabolas). Also, sums of convex functions are convex (this is the reason why ordinary least squares is a convex problem), so we can focus on making each term convex separately.

For this, we first find the minimum and maximum values of each residual $r_i$ is over the given interval, denoted as $r_{min}^i$ and $r_{max}^i$ respectively. The key geometric idea of our relaxation is to reweight these residuals such that after reweighting, their maximum becomes $\epsilon^2$ and thus they are not affected by truncation (Fig. \ref{fig:wlsrelaxideawithrelax}).

However, some additional special case handling is required to ensure that this relaxation becomes tight when computing over (small) intervals where the TLS objective is already convex. This is crucial to ensure that BnB converges in a finite number of iterations. First, the reweighting should only be applied to residuals where $r_{min} < \epsilon^2$, i.e., residuals that are not already outliers over the entire interval. Similarly, residuals where $r_{max} < \epsilon^2$ are inliers over the entire interval and are therefore already convex and do not need to be reweighted.

After considering these special cases, we can show that $\text{WLS}_{r}$ is a convex relaxation of \ref{eq:tls-scalar-for-wls} over a given interval:
\begin{equation}
	\label{eq:tls-wls-relaxation1}
	\begin{aligned}	
		\text{WLS}_{r} &= 
		\sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right) + \sum_{i=1}^{N} o_i\epsilon^2\\
	\end{aligned}
\end{equation}

with the weights $w_i$ defined as:
\begin{equation}
	\begin{aligned}	
		w_i  = 
		\begin{cases}
			\frac{\epsilon^2 - r_{min}^i}{r_{max}^i - r_{min}^i} &  r_{max}^i > \epsilon^2\\
			1 & \, \text{otherwise}
		\end{cases}
	\end{aligned}
\end{equation}

and the binary variables $o_i$ deciding whether the $i$-th residual is an outlier or not:
\begin{equation}
	\begin{aligned}	
		o_i &= \begin{cases}
			1 &  r_{min}^i > \epsilon^2\\
			0 & \, \text{otherwise}
		\end{cases}
	\end{aligned}
\end{equation}

For a proof of this WLS-relaxation, see \ref{proof:tls-convex-wls-relaxation}.

\subsection{Solving the equality-constrained case}

If the inequality-constraint is not satisfied after the first unconstrained solve, we need to solve the problem with equality-constraint, that is the following optimization problem:

\begin{equation}
	\label{eq:tls-rwls-eq}
	\begin{aligned}
		\argmin_{\Delta \mathbf{R} \in \SOthree}  \quad &\sum_{i=1}^{N} w_i ||\mathbf{R}_c^\transposed \mathbf{a}_i - \Delta\mathbf{R} \mathbf{b}_i ||_2^2 \\
		\text{subject to} \quad &d_{\angle}(\mathbf{I}, \Delta\mathbf{R}) = n_r
	\end{aligned}
\end{equation}

We can show that this optimization problem is equivalent to the following problem that finds the optimal axis of rotation $\mathbf{n}^*$:

\begin{equation}
	\begin{aligned}
		\label{eq:eq-constrained-prob-ev}
		\mathbf{n}^* = \argmin_{\mathbf{n} \in \Rthree} \quad &-\left(\mathbf{n}^\transposed \mathbf{A} \mathbf{n} + 2 \mathbf{g}^\transposed \mathbf{n} \right)\\
		\text{subject to} \quad &|| \mathbf{n}|| = 1\\
	\end{aligned}
\end{equation}

with the data matrices $\mathbf{A}$ and $\mathbf{g}$ that depend on the points $\mathbf{a}_i$ and $\mathbf{b}_i$ and the angle constraint $n_r$:

\begin{equation}
	\label{eq:davenport-data-matrices}
	\begin{aligned}
		\mathbf{B} &= \sum_{i=1}^N w_i \mathbf{R}_c^\transposed \mathbf{a}_i\mathbf{b}_i^\transposed \in \Rone^{3 \times 3}, \quad
		\mathbf{z} = \sum_{i=1}^N w_i [\mathbf{a}_i]_{\times}\mathbf{b}_i \in \Rone^{3}, \quad \mathbf{C} = \mathbf{B} + \mathbf{B}^\transposed\\
		\mathbf{A} &= \sin^2 \left(\frac{n_r}{2} \right) \mathbf{C}, \quad \mathbf{g} =  \sin \left(\frac{n_r}{2} \right) \cos \left(\frac{n_r}{2} \right) \mathbf{z}
	\end{aligned}
\end{equation}

This derivation is based on the Daveport's Quaternion-method \cite{davenportsQ} \cite{8594296}, see \ref{proof:wls-relax-eq-constrained-rotation} for a proof of the equivalence of both optimization problems.

To solve the optimization problem \ref{eq:eq-constrained-prob-ev}, we perform overall two steps: We first need to compute the eigenvalue decomposition  $\mathbf{A} = \mathbf{Q}\mathbf{\Lambda}\mathbf{Q}^\transposed$ and then need to find largest root of the following rational function $C(\gamma)$ (characteristic polynomial):

\begin{equation}
	\label{eq:characterisitc-poly}
	\begin{aligned}
		C(\gamma) = \sum_{i=1}^3 \left(\frac{a_i}{\lambda_i - \gamma}\right)^2 - 1
	\end{aligned}
\end{equation}

with
\begin{equation}
	\begin{aligned}
		\mathbf{a} = (a_1, a_2, a_3)^\transposed = \mathbf{Q}^\transposed \mathbf{g} \in \Rthree, \quad \mathbf{\Lambda} = \text{diag}(\lambda_1, ..., \lambda_3)
	\end{aligned}
\end{equation}

\begin{algorithm}[!ht]
	\caption{Solve constrained eigenvalue problem}
	\label{alg:solve-constrained-ev}
	\begin{algorithmic}[1]
		\Require{Data matrices $\mathbf{A} \in \Rone^{3\times3}, \mathbf{g} \in \Rthree$}
		\Ensure Solution $\mathbf{n}^*$ to the optimization problem Eq. \ref{eq:eq-constrained-prob-ev}
		\hypertarget{ref:min-constrained-ev}{
			\Function{\code{solve\_constrained\_ev}}{$\mathbf{A}, \mathbf{g}$}}
		\State $\code{kEps} \leftarrow 2^{-48}$ \algorithmiccomment{Epsilon-value for double precision}
		\State $\mathbf{Q}, \mathbf{\Lambda} \leftarrow \code{eig(} \mathbf{A} \code{)}$ \algorithmiccomment{Compute eigen-decomposition}
		\State $k = \argmax\{\lambda_1, \lambda_2, \lambda_3\}$
		\State $\mathbf{a} \leftarrow \mathbf{Q}^\transposed \mathbf{g}$
		\State $l = \lambda_k + |a_k| - \code{kEps}, r = ||\mathbf{a}|| + \lambda_k + \code{kEps}$ \algorithmiccomment{Bracket root}
		\State $\hat{\gamma} \leftarrow \code{bisect} (C, [l, r] \code{)}$ \algorithmiccomment{Find root of Eq.\ref{eq:characterisitc-poly} with bisection}
		\State $\mathbf{f} \leftarrow \left(\frac{a_1}{\lambda_1 - \hat{\gamma}}, \frac{a_2}{\lambda_2 - \hat{\gamma}}, \frac{a_3}{\lambda_3 - \hat{\gamma}} \right)$
		\State $\mathbf{n}^* \leftarrow \mathbf{Q}\mathbf{f}$ 
		\State \Return $\mathbf{n}^*$
		\EndFunction
	\end{algorithmic}
\end{algorithm}

This algorithm is shown in Alg \ref{alg:solve-constrained-ev}, see \ref{sec:constrained-ev-problem-derivation} for a derivation with explanations.
The eigen-decomposition can exploit the fact that the matrix is real and symmetric (but not positive semi-definite) for which faster algorithms are available (i.e. Eigen's \code{Eigen::SelfAdjointEigenSolver}). Note also that the vector $\mathbf{f}$ in line 8 is equal to $(\mathbf{\Lambda} - \hat{\gamma} \mathbf{I})^{-1} \mathbf{a}$. No actual matrix-inversion is required however since the matrix is diagonal and also with floating-point arithmetic it is numerically more stable/accurate to compute $a/b$ instead of $\frac{1}{b} a$.

\subsection{Overall active-set solver}

\begin{algorithm}[!ht]
	\caption{Active-set method for minimizing the WLS-relaxation over rotations (\ref{eq:tls-rwls})}
	\label{alg:min-wls-relax-rot}
	\begin{algorithmic}[1]
		\Require{Point cloud matrices $\mathbf{A}, \mathbf{B}$, node interval $(\mathbf{R_c}, n_r)$, computed WLS-relaxation weights vector $\mathbf{w}$}
		\Ensure Solution to the optimization problem \ref{eq:tls-rwls}
		\hypertarget{ref:min-wls-relax-rot2}{
			\Function{\code{Minimize\_WLS\_Relax\_Rotation}}{$\mathbf{A}, \mathbf{B}, \mathbf{R_c}, n_r, \mathbf{w}$}}
		\State $\mathbf{B} \leftarrow \sum_{i=1}^N w_i \mathbf{R}_c^\transposed \mathbf{a}_i\mathbf{b}_i^\transposed $ \algorithmiccomment{Compute cross-correlation matrix}
		\State $\bar{\Delta \mathbf{R}} \leftarrow \code{svd\_rot(} \mathbf{B} \code{)}$  \algorithmiccomment{Solve unconstrained with SVD-algorithm}
		\If{$d_{\angle}(\mathbf{I}, \Delta\mathbf{R}) \leq n_r$} \algorithmiccomment{Check if constraint is already satisfied}
		\State \Return $\mathbf{R_c} \bar{\Delta \mathbf{R}}$
		\EndIf
		\State $\mathbf{A}, \mathbf{g} \leftarrow \code{get\_davenport\_matrices(} \mathbf{B}, n_r \code{)}$ \algorithmiccomment{Get  data-matrices (Eq. \ref{eq:davenport-data-matrices})}
		\State $\mathbf{n}^* \leftarrow \code{solve\_constrained\_ev(} \mathbf{A}, \mathbf{g} \code{)}$ \algorithmiccomment{Solve for
			rotation axis (Eq. \ref{eq:eq-constrained-prob-ev})}
		\State $\bar{\Delta \mathbf{R}} \leftarrow \code{from\_axis\_angle(}\mathbf{n}^*, n_r \code{)}$ \algorithmiccomment{ Construct optimal rotation}
		\State \Return $\mathbf{R_c} \bar{\Delta \mathbf{R}}$
		\EndFunction
	\end{algorithmic}
\end{algorithm}

Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is shown in Alg. \ref{alg:min-wls-relax-rot}. It is very efficient since the only computationally intensive operation is the computation of the $3 \times 3$ cross-correlation matrix $\mathbf{B}$ (line 13) that has linear runtime in the number of points. All the operations that follow are based on this $3 \times 3$ matrix and therefore have a constant runtime (i.e. independent of the number of points) and are in practice very fast. For example, the eigen-decomposition (line 3) takes only around $10 \mu s$ while the computation of the cross-correlation matrix requires aroundd $500 \mu s$ for 10k points. This is also because the cross-correlation matrix is re-used between the SVD-algorithm (line 3) and the algorithm for the equality-constraint case (line 7). The function \code{from\_axis\_angle} converts a given axis and angle to a rotation matrix using Rodrigues' formula.

This solver is novel to our knowledge -- the problem of solving the Wahba-problem with inequality-constrained angle has not been proposed in the literature before.

The solver for the equality-constraint case ($\code{solve\_constrained\_ev}$) is implemented from the derivation first done in \cite{10.1007/978-3-642-75536-1_57} and inspired by the \textit{Trust-Region Subproblem} (TRS) that arises in trust-region (local) optimization algorithms \cite{Adachi2017SolvingTT} such as Levenberg-Marquardt. For such TRS-problems, an active-set method has been proposed also in \cite{Rontsis2022} where it was generalized additionally for a minimum-norm constraints.

### 2025-04-21 sec/4_evaluation.tex
Started convex relaxation based solver (MA thesis) paper

\section{Experiments}
In this section, we present experiments where we study the intersection graphs arising from the robust point cloud registration problem. We evaluate the runtime of state-of-the-art maximum clique finding algorithms.

\label{sec:evaluation}
We perform the experiments on a desktop PC with an Intel i7 12700KF CPU and 32GiB of RAM.
\subsection{Synthetic experiments}

\subsection{Synthetic experiments}
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/execution_time_over_outlier_rate_boxplot}
	\includegraphics[width=0.49\linewidth]{figures/rotation_error_over_outlier_rate.pdf}
	\caption{Execution time and rotation error over different outlier rates with $N=100$ on synthetic data.}
	\label{fig:execution-time-over-outlier-rate}
\end{figure}

We generate synthetic point clouds by sampling inliers from a unit cube $s \cdot [-1, 1]^3$ with scale $s$. We add uniform measurement noise by sampling a ball of radius $\epsilon^2$, and then sample outliers in the bounding box of the inliers, also uniformly. First, we set $s=10$ and $\epsilon=0.5$. As shown in Fig. \ref{fig:execution-time-over-outlier-rate}, the runtime is very low for $N =100$ and decreases until 90\% outlier-rate because the graph sparsity increases. Above 90\% outlier-rate, it increases because the number of visited cliques increases.

\subsubsection{Scalability}
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/scalability-synth}
	\caption{Scalability on synthetic data, outlier rate $75\%$}
	\label{fig:execution-time-over-n}
\end{figure}

We evaluate the scalability on synthetic data at $75\%$ outlier-rate, results are shown in Fig \ref{fig:execution-time-over-n}. Our solves is efficient since it implicitly prunes vertices based on the TLS objective.

\subsubsection{Validity of the global optimality certificate}
Even though GNC is a heuristic, based on our case analysis (Sec. \ref{sec:branching-over-comb}) it seems likely that GNC finds global minimum under certain conditions.
We therefore test on synthetic data whether the global optimality certificate is valid, i.e. whether at termination, the lower bound is less or equal than the TLS objective at the ground-truth solution.
We define the suboptimality $\eta$ of the solution based on the upper bound (UB), that is, the best known solution, and the lower bound (LB) \cite{9785843}
\begin{equation}
	\begin{aligned}    
		\eta = \frac{\text{UB} - \text{LB}}{1 + |\text{UB}| + |\text{LB}|}
	\end{aligned}
\end{equation}
We test with $N=100$, $\epsilon=0.5$ and vary the data scale $s$ from 3 to 100. We test for each scale outlier rates of $0\%, 30\%, 50\%, 90\%, 96\%$, and do 1000 trials for each combination.
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/certificate_validity_vs_scale}
	\includegraphics[width=0.49\linewidth]{figures/eta_subopt_over_outlier_rate}
	\caption{Amount of instances with valid global optimality certificate (i.e. valid suboptimality) on synthetic data, including achieved $\eta$-suboptimality.}
	\label{fig:cert-validity}
\end{figure}
The results are shown in Fig. \ref{fig:cert-validity}, we clearly see that with a larger data scale, the likelihood of a valid certificate (and successful registration result) increases. While this observation alone could be explained solely by the solver being able to solve the problem more likely, we also see that the suboptimality increases for instances above 90\% outliers. This indicates that our solver is able to detect that some instances could not be solved to global optimality.
Unfortunately, being only able to reject some and not all instances means our solver is currently neither globally optimal (being able to find the global minimum of every instance and outputting a valid lower bound on every instance), nor certifiable (outputting only a valid lower bound on every instance).

\subsubsection{Comparison with SDP-relaxation solver STRIDE} Currently the only solver that can solve the TLS registration problem of size $N=100$ to proven global optimality is STRIDE \cite{9785843}. We compare it on our synthetic data to access the achieved suboptimality and estimation errors. We use the public MATLAB-implementation of the authors and use the MOSEK solver for the chordal initialization of STRIDE.
We set $\epsilon = 0.5$, $N=30$ and perform three trials per outlier rate.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/subopt_stride}
	\includegraphics[width=0.49\linewidth]{figures/runtime_stride}
	\caption{The reached suboptimality and the runtime using STRIDE over the outlier rate on synthetic data, $N=30$.}
	\label{fig:stride-subopt-eval}
\end{figure}
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/rotation_error_stride}
	\caption{Rotation error using STRIDE.}
	\label{fig:stride-rot-err}
\end{figure}

The results are shown in Fig. \ref{fig:stride-subopt-eval}, STRIDE is able to find and certify the global optimum as long as the outlier rate is below 90\%. The runtime of STRIDE is even with as few as 30 points in the order of $10^3$ seconds. STRIDE achieves a similar rotation error compared to our solver in instances that could be solved (Fig. \ref{fig:stride-rot-err}). Note that the rotation error is meaningless for instances that STRIDE could not solve (i.e. above 80\%), since it is always able to detect them.

\subsection{Results on 3DMatch dataset}
\input{sec/eval-3DMatch-table}
We evaluate our solver on the 3DMatch dataset \cite{zeng20163dmatch} 
following \cite{SC2-PCR-Chen-2022-CVPR, zhang20233d} on all scenes. We evaluate the residual rotation (RE) and translation (TE) and the registration recall (RR). A registration is defined as successful if RE $\leq 15$ degrees and TE $\leq 30$ cm. We sub-sample the correspondeces to 500 and set the inlier-threshold to $\epsilon = 0.6$. For the other methods, we report the results from \cite{zhang20233d} since the test setup is identical. The results are shown in Tab. \ref{tab:3DMatch-results}.

\subsection{Results on KITTI dataset}
\input{sec/eval-kitti-table}

We evaluate our method on the KITTI odometry dataset \cite{Kitti-Geiger2013IJRR} following \cite{SC2-PCR-Chen-2022-CVPR, zhang20233d} and similar to the 3DMatch dataset. A registration is defined as successful if RE $\leq 5$ degrees and TE $\leq 60$ cm.
We sub-sample the correspondeces to 500 and set the inlier-threshold to $\epsilon = 0.6$. We set a time limit of 100 milliseconds. The FCGF \cite{9009829} descriptor has on average an outlier-rate of 75\%, while the FPFH has usually above 90\% outlier-rate. For the other methods, we report the results from \cite{zhang20233d} since the test setup is identical. The results are shown in Tab. \ref{tab:eval-kitti}, our method achieves comparable results to other state-of-the-art methods without dataset-specific tuning of parameters, outperformed only by MAC and SC$^2$-PCR.

### 2025-04-21 sec/5_conclusion.tex
Started convex relaxation based solver (MA thesis) paper

\section{Conclusion}
\label{sec:conclusion}

In this paper, we presented a general approach of minimizing the Truncated Least Squares (TLS) objective motivated by the geometry of the problem. We showed how intersection graphs allow for an elegant formulation of the combinatorial problem and how the global minimum can be found among the maximal cliques. We showed that surprisingly, the previously studied compatibility graphs motivated by problem-specific geometric observations lead to these special classes of graphs, namely the intersection graphs. 
We give a sufficient condition under which the global optimum is a maximal clique and discuss cases where subcliques may be the global minimum.
Unfortunately, we were not able to fully prove that we can detect all conditions under which subcliques may be the global optimum.
We think that a better understanding of the $\SEthree$-manifolds and their geometry including their combinatorial properties could lead to efficient algorithms for detecting such cases.

### 2025-04-21 sec/X_suppl.tex
Started convex relaxation based solver (MA thesis) paper

\clearpage
\setcounter{page}{1}
\maketitlesupplementary

\begin{theorem}[\textbf{Knowing the ground-truth inliers solves the TLS-problem}]
	Given the ground-truth inlier set $\mathcal{I}^*$, the global optimum $(\MR^*, \vt^*)$ is found by solving the least-squares problem:
	\begin{equation}
		\begin{aligned}
			(\MR^*, \vt^*) = \argmin_{(\MR, \vt) \in \SEthree} \sum_{i \in \mathcal{I}^*} \normsq{\MR \vp_i - \vq_i + \vt}
		\end{aligned}
	\end{equation}
\end{theorem}

\begin{proof}
    \label{proof:gt-corrs-solves-tls}
	Since we know the ground-truth inlier set, we know which residuals are less or equal $\epssq$ and therefore not truncated. Since $\min(x, \epssq) = x$ for all $x \leq \epssq$, we can equivalently rewrite the TLS-objective \ref{eq:pcr-tls} as:
    \begin{equation}
	\begin{aligned}
		&\sum_{i=1}^{N} \min \left(\normsq{\MR \vp_i - \vq_i + \vt}, \epsilon^2 \right)\\
        &= \sum_{i \in \mathcal{I}^*} \normsq{\MR \vp_i - \vq_i  + \vt}  + \sum_{i \in \{1, \cdots, N\}  \setminus \mathcal{I}^*} \epsilon^2
	\end{aligned}
    \end{equation}
    Since the second sum is constant, it does not affect the minimizer of the least-squares term (the first sum). Therefore, minimizing only  the first, i.e. solving the least-squares problem solves the TLS-problem.
\end{proof}

\begin{lemma}[\textbf{Minimum of rotation residual}]\label{lemma:ls-rot}
	The minimum for all $\va, \vb \in \Rthree$
	\begin{equation}
		\begin{aligned}
			&\min_{\MR \in \SOthree} \normsq{\MR \va - \vb}\\
			&= \left( \norm{\va} - \norm{\vb} \right)^2
		\end{aligned}
	\end{equation}
\end{lemma}

\begin{proof}
\begin{equation}
	\begin{aligned}
		&\min_{\MR \in \SOthree} \normsq{\MR \va - \vb}\\
		=&\min_{\MR \in \SOthree} \left( \MR \va - \vb \right)^\transposed \left( \MR \va - \vb \right)\\
		=&\min_{\MR \in \SOthree} (\MR\va)^\transposed \MR\va - 2 (\MR \va)^\transposed \vb + \vb^\transposed \vb\\
		=&\min_{\MR \in \SOthree} \va^\transposed \MR^\transposed \MR \va - 2 (\MR \va)^\transposed \vb + \vb^\transposed \vb\\
		=&\min_{\MR \in \SOthree} \va^\transposed \va - 2 (\MR \va)^\transposed \vb + \vb^\transposed \vb\\
		=&\min_{\MR \in \SOthree} \normsq{\va} - 2 (\MR \va)^\transposed \vb  + \normsq{\vb}\\
		=& \normsq{\va} - 2 \norm{\va} \norm{\vb}  + \normsq{\vb}\\
		=& \left( \norm{\va} - \norm{\vb} \right)^2
	\end{aligned}
\end{equation}

\end{proof}

\begin{lemma}[\textbf{Least-squares solution with two points}]\label{lemma:ls-two-points}
	In the special case of two points, the least squares solution for point cloud registration is
	\begin{equation}
		\begin{aligned}
			\min_{(\MR, \vt) \in \SEthree} & \normsq{\MR \vp_i - \vq_i  + \vt} +  \normsq{\MR \vp_j - \vq_j  + \vt}\\
			&= \frac{1}{2} \left( \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} \right)^2
		\end{aligned}
	\end{equation}
	Both residual terms are equal at the minimum.
\end{lemma}

\begin{proof}
	A well known result is that the optimal translation is related to the optimal rotation and the centroids of both point clouds (see e.g \cite{Kabsch-1976-Point-set-alignment, sorkine2017least}):
	\begin{equation}
		\begin{aligned}
			\vt &= \bar{\vq} - \MR \bar{\vp}\\
			\bar{\vp} &= \frac{\vp_i + \vp_j}{2}, \, \bar{\vq} = \frac{\vq_i + \vq_j}{2}
		\end{aligned}
	\end{equation}
	We insert the optimal translation into the objective and simplify:
	\begin{equation}\label{eq:insert-t-opt}
		\begin{aligned}
			&\normsq{\MR \vp_i - \vq_i  + (\bar{\vq} - \MR \bar{\vp})} +  \normsq{\MR \vp_j - \vq_j  + (\bar{\vq} - \MR \bar{\vp})}\\
			=&\normsq{\MR (\vp_i - \bar{\vp}) - (\vq_i -\bar{\vq})} + \normsq{\MR (\vp_j - \bar{\vp}) - (\vq_j -\bar{\vq})} \\
			=&\normsq{\MR \left(\vp_i - \frac{\vp_i + \vp_j}{2} \right) - \left(\vq_i - \frac{\vq_i + \vq_j}{2} \right)}\\
			+ &\normsq{\MR \left(\vp_j - \frac{\vp_i + \vp_j}{2} \right) - \left(\vq_j -\frac{\vq_i + \vq_j}{2} \right)} \\
			=&\normsq{ \frac{1}{2}(\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j))}\\
			+ &\normsq{ \frac{1}{2}(\MR (\vp_j - \vp_i ) - (\vq_j - \vq_i))}\\
			=& 2\normsq{ \frac{1}{2}(\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j))}\\
			=& \frac{1}{2}\normsq{\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j)}\\
		\end{aligned}
	\end{equation}
	We just showed that both residual terms are equal at the optimal translation.	
	What remains is to optimize over the rotation $\MR$. By lemma \ref{lemma:ls-rot}, the minimum over rotations is:
	\begin{equation}
		\begin{aligned}
			&\min_{(\MR, \vt) \in \SEthree} \normsq{\MR \vp_i - \vq_i  + \vt} +  \normsq{\MR \vp_j - \vq_j  + \vt}\\
			&=\min_{\MR \in \SOthree} \frac{1}{2}\normsq{\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j)}\\
			&= \frac{1}{2} \left( \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} \right)^2
		\end{aligned}
	\end{equation}
\end{proof}

\begin{theorem}[\textbf{Non-empty intersection of T-Sets is equivalent to compatible TIMs}]
	\begin{equation}
		\begin{aligned}
			&\mathcal{T}_i \cap \mathcal{T}_j \neq \emptyset\\
			\iff &\exists (\MR, \vt) \in \SEthree:\\
			&\norm{\MR \vp_i - \vq_i + \vt} \leq \epsilon\\
			\land &\norm{\MR \vp_j - \vq_j + \vt} \leq \epsilon\\
			\iff &| \norm{\va_{ij}} - \norm{\vb_{ij}} | \leq 2 \epsilon
		\end{aligned}
	\end{equation}
	where 
	\begin{equation}
		\label{eq:tims}
		\begin{aligned}
			\va_{ij} &= \vp_i - \vp_j\\
			\vb_{ij} &= \vq_i - \vq_j
		\end{aligned}
	\end{equation}
\end{theorem}

\begin{proof}
	\label{proof:non-empty-intersection-implies-tims-compat}
	Whether there exists a $(\MR, \vt) \in \SEthree$ such that $\norm{\MR \vp_i - \vq_i  + \vt} \leq \epsilon$ is equivalent to whether the minimum is $\leq \epsilon$, i.e. whether $\min_{(\MR, \vt) \in \SEthree} \norm{\MR \vp_i - \vq_i  + \vt} \leq \epsilon$ holds.
	By lemma \ref{lemma:ls-two-points} by minimizing the sum of two residuals, at the minimum they are both equal and therefore if their sum is $\leq 2\epssq$, they both are $\leq \epssq$. We conclude therefore
	\begin{equation}
		\begin{aligned}
			&\mathcal{T}_i \cap \mathcal{T}_j \neq \emptyset \iff\\
			&\min_{(\MR, \vt) \in \SEthree} \normsq{\MR \vp_i - \vq_i  + \vt} + \normsq{\MR \vp_j - \vq_j  + \vt} \leq 2\epssq\\
			& \iff \frac{1}{2} \left( \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} \right)^2 \leq 2\epssq\\
			& \iff | \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} | \leq 2\epsilon\\
			& \iff | \norm{\va_{ij}} - \norm{\vb_{ij}} | \leq 2 \epsilon
		\end{aligned}
	\end{equation}
\end{proof}

\begin{lemma}[\textbf{Consensus sets are cliques in the intersection graph}]
	Consensus sets are cliques in the intersection graph.
\end{lemma}

\begin{proof}
	Cliques in the intersection imply by definition pairwise non-empty intersection of T-Sets. Since pairwise non-empty intersection is a necessary condition for total non-empty intersection and therefore consensus sets, consensus sets are cliques in the intersection graph.
\end{proof}

\begin{lemma}[\textbf{Maximal clique enumeration and global optimality}]
	\label{lemma:max-clique-non-helly-app}
	Maximal clique enumeration finds the ground-truth consensus set and therefore the global minimum of an instance if all maximal cliques of this instance are maximal consensus sets.
\end{lemma}

\begin{proof}
	The key statement of the lemma is that if every maximal clique is a maximal consensus set, then subcliques of maximal cliques (which by definition are not maximal cliques) cannot be the ground truth inlier set. Consequently, they need not be enumerated.

	The global optimum lies in the union of all T-sets, i.e. $\bigcup_{i \in \RangeOneToN} \mathcal{T}_i$. Every vertex of a graph is contained in at least one maximal clique of that graph. Therefore, the global optimum is contained in at least one union $\bigcup_{i \in \vc} \mathcal{T}_i$ of a maximal clique $\vc$.

	The intersection of the T-sets of any subclique contains the intersection of the T-sets of a maximal clique containing that subclique. Therefore, if the consensus set is maximal, the best solution (minimizing the TLS objective) that can be found among subcliques is no better than this maximal clique. If this holds for every maximal clique of an instance, then the ground-truth inlier set is not a subclique of any maximal clique, and therefore must itself be a maximal clique. Thus, enumerating maximal cliques will find the ground-truth consensus set if all maximal cliques of this instance are maximal consensus sets. v
\end{proof}

\label{proof:wls-interval-analysis-rotation}

\begin{lemma}[\textbf{Consensus set size and TLS objective }]
	Given an upper-bound $\text{UB}$ on the TLS objective, the ground-truth consensus set must have at least $d_{m}$:
    \begin{equation}
        \begin{aligned}
            d_m = \lceil N - \text{UB} / \epssq \rceil
        \end{aligned}
    \end{equation}
\end{lemma}

\begin{proof}
    By the definition of the TLS objective, 
    $r_{in} + (N - d_m) \epsilon^2 \leq \text{UB}$ must hold, where $r_{in}$ is the (unknown) sum of the inlier residuals. Rearranging yields $d_m \geq r_{in}  + N - \text{UB} / \epssq$.
    Since all the summands in the TLS objective are non-negative, $r_{in} = 0$ is a valid lower bound. Since the number of elements in the consensus set $d_m$ must be an integer, we conclude that $d_m = \lceil N - \text{UB} / \epssq \rceil$.
\end{proof}

### 2025-04-21 sec/dim_decomp.tex
Started convex relaxation based solver (MA thesis) paper

\subsection{Dimensionality decomposition: globally optimal or not ?}
We have described so far only how to solve the rotation problem using the TLS objective. To solve for the full 6 Degrees of Freedom (DoF), i.e. both rotation and translation, we apply the decomposition technique introduced in Yang et al. (TEASER \cite{Yang20tro-teaser}). They popularized the idea of first removing the translation and then solving sequentially for rotation and translation. This idea works well in practice and spawned recently many other works using the same idea or variations of it, e.g. \cite{Yang20tro-teaser} \cite{10432947} and \cite{9878458} \cite{9879801}. However, Yang et al. do not state whether this decomposition preserved optimality -- they do not comment on the equivalence of both optimization problems. It was recently hinted that this idea may break the global optimalit, but with no counter-examples, a final anwswer was unknown.

We give a definite anwswer to this -- it unfortunatelly breaks global optimality. 
On difficult instances, we generated many counter-examples where solving sequentially for rotation and translation does not yield the overall globally optimal 3D-transformation (see sec TODO).

At the same time, we are able to "save" the idea that works well in practice by discovering how the rotation-subproblem is related to the original registration problem: Although it is not equivalent, the TIMS-Rot problem provides a lower bound on the TLS-Reg problem.

This gives a valuable insight into why heuristic methods using this idea work so well in practice (TODO global optimum is close or what?). 
We output this sub-optimality explicitly and therefore obtain a certifiable algorithm (TODO yeah...)\\

### 2025-04-21 sec/downprojection_lb.tex
Started convex relaxation based solver (MA thesis) paper

\textbf{Down-projecting the lower bound from TIMS-Rot}. After solving the TIMS-Rot problem, we can obtain from it a lower bound on the original registration problem (Eq. \ref{eq:pcr-tls}). We first obtain from (already minimized) WLS relaxation the number of potential inliers $N_{in}$ and the inlier score value $W_{in}$:

\begin{equation}
	\label{eq:wls-relax-info}
	\begin{aligned}
		N_{in} &:= \sum_{i=1}^{N} (1 - o_i)\\
		W_{in} &:= \sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right)
	\end{aligned}
\end{equation}

TODO prove. after theorem 6 turns out to be wrong, do we need to alter this formula? like do we already count each vertex touched by at least one edge?\\
\begin{equation}
	\label{eq:lb-downprojection}
	\begin{aligned}
		R_{in} &= \sqrt{\frac{2 N_{in} + N}{N^2}}\\
		TLS_{in} &= \frac{W_{in}}{N \cdot R_{in}}\\
		TLS_{out} &= (1 - R_{in}) \epsilon^2\\
		project-lb  &= tls_{in} + tls_{out}
	\end{aligned}
\end{equation}

### 2025-04-21 sec/eval-kitti-table.tex
Started convex relaxation based solver (MA thesis) paper

\begin{table}[!ht]
    \centering
    \caption{Registration accuracy on KITTI dataset in comparison with other methods:\label{tab:eval-kitti}}
    \resizebox{\linewidth}{!}{
      \begin{tabular}{l|rrr|rrr}
      \hline
      \multirow{2}[2]{*}{} Descriptor& \multicolumn{3}{c|}{FPFH} & \multicolumn{3}{c}{FCGF} \\
            & \multicolumn{1}{l}{RR(\%)} & \multicolumn{1}{l}{RE(\textdegree)} & \multicolumn{1}{l|}{TE(cm)} & \multicolumn{1}{l}{RR(\%)} & \multicolumn{1}{l}{RE(\textdegree)} & \multicolumn{1}{l}{TE(cm)} \\
      \hline
      TEASER++\cite{Yang20tro-teaser}   &  91.17  &  1.03  & 17.98 & 94.96   &  0.38  & 13.69 \\
      RANSAC  &  74.41  &  1.55  &  30.20 & 80.36  &  0.73  & 26.79 \\
      CG-SAC\cite{quan2020compatibility}  & 74.23  & 0.73  & 14.02 & 83.24  & 0.56  & 22.96 \\
      $\rm{SC}^2$-PCR\cite{SC2-PCR-Chen-2022-CVPR} & 99.64  &  0.32  & 7.23  & 98.20 & 0.33  & 20.95 \\
      MAC \cite{zhang20233d} &  99.46  &  0.4  & 8.4  &  97.84 &  0.34  & 19.34 \\
      \hline
      \MyAlgo{} (Ours) &  95.50 & 0.77  & 16.43  &  96.76 & 0.34 & 20.59 \\
      \hline
      \end{tabular}}
  \end{table}%

### 2025-04-21 sec/eval-table.tex
Started convex relaxation based solver (MA thesis) paper

\begin{table*}[!ht]
	\centering
	\scriptsize
	\renewcommand\tabcolsep{4.0pt}
	\begin{tabular}{cc|cccccc|cccccc|c}%\item The \LaTeX{}-generated files (e.g. .aux and .bbl file, etc.) if they are needed to compile your source (generally, the .aux and bbl file can be omitted; the bib file might be necessary, however).

		\hline%\end{itemize}
		& & \multicolumn{6}{c|}{FPFH (traditional descriptor)} & \multicolumn{6}{c|}{FCGF (learning based descriptor)} \\ 
		& & RR(\%$\uparrow$) & RE($\circ$$\downarrow$) & TE(cm$\downarrow$) & IP($\%$$\uparrow$) &  IR($\%$$\uparrow$) & F1($\%$$\uparrow$) & RR($\%$$\uparrow$) & RE($\circ$$\downarrow$) & TE(cm$\downarrow$) & IP($\%$$\uparrow$) &  IR($\%$$\uparrow$) & F1($\%$$\uparrow$) & Time(s) \\ \hline
		\multirow{7}{0.04cm}{\rotatebox{90}{learning}} & DCP* \cite{wang2019deep} & 3.22 & 8.42 & 21.40  & - & - & - & 3.22 & 8.42 & 21.40  & - & - & - & 0.07\\
		& PointNetLK* \cite{aoki2019pointnetlk} & 1.61 & 8.04  & 21.30 & - & - & - & 1.61 & 8.04 & 21.30 & - & - & - & 0.12 \\
		& 3DRegNet \cite{pais20203dregnet} & 26.31 & 3.75 & 9.60 & 28.21 & 8.90 & 11.63 & 77.76 & 2.74 & 8.13 & 67.34 & 56.28 & 58.33 & \textcolor{Blue}{0.05}\\
		& DGR \cite{choy2020deep} & 32.84 & 2.45 & 7.53 & 29.51 & 16.78 & 21.35 & 88.85 & 2.28 & 7.02 & 68.51 & 79.92 & 73.15 & 1.53\\

		& DHVR \cite{lee2021deep} & 67.10 & 2.78 & 7.84 & 60.19 & 64.90 & 62.11 & 91.93 & \textcolor{Blue}{2.25} & 7.08 & \textcolor{Blue}{80.20} & 78.15 & 78.98 & 3.92\\
		& DHVR-Origin \cite{lee2021deep} & \textcolor{Blue}{80.22} & \textcolor{Blue}{2.06} & 6.87 & & & & 91.40 & \textcolor{Red}{2.08} & 6.61 & & & & 0.46\\
		& PointDSC \cite{bai2021pointdsc} & 77.57 & \textcolor{Red}{2.03} & \textcolor{Red}{6.38} & 68.45 & \textcolor{Blue}{71.56} & \textcolor{Blue}{69.75} & \textcolor{Blue}{92.85} & \textcolor{Red}{2.08} & \textcolor{Red}{6.51} & 78.91 & 86.23 & \textcolor{Blue}{82.12} & 0.10\\ \hline
		% 		& PointDSC-gen \cite{bai2021pointdsc} & 68.12 & 2.69 & 7.91 & \textcolor{Red}{83.63} & 20.06 & 31.12 & 87.74 & \textcolor{Red}{1.76} & \textcolor{Red}{6.02} & \textcolor{Red}{95.77} & 42.39 & 57.19 & 0.10\\  \hline

		\multirow{10}{0.04cm}{\rotatebox{90}{traditional}} & SM \cite{leordeanu2005spectral} & 55.88 & 2.94 & 8.15 & 47.96 & 70.69 & 50.70 & 86.57 & 2.29 & 7.07 & 81.44 & 38.36 & 48.21 & \textcolor{Red}{0.03}\\
		& ICP* \cite{besl1992method} & 5.79 & 7.93 & 17.59 & - & - & - & 5.79 & 7.93 & 17.59 & - & - & - & 0.25\\
		& FGR \cite{zhou2016fast} & 40.91 & 4.96 & 10.25 & 6.84 & 38.90 & 11.23 & 78.93 & 2.90 & 8.41 & 25.63 & 53.90 & 33.58 & 0.89 \\
		& TEASER \cite{yang2020teaser} & 75.48 & 2.48 & 7.31 & \textcolor{Red}{73.01} & 62.63 & 66.93 & 85.77 & 2.73 & 8.66 & \textcolor{Red}{82.43} & 68.08 & 73.96 & 0.07\\
		& GC-RANSAC \cite{barath2018graph} & 67.65 & 2.33 & 6.87 & 48.55 & 69.38 & 56.78 & 92.05 & 2.33 & 7.11 & 64.46 & \textcolor{Red}{93.39} & 75.69 & 0.55\\
		%& RANSAC-1k & 40.05 & 5.16 & 13.65 & 51.52 & 34.31 & 39.23 & 86.57 & 3.16 & 9.67 & 76.86 & 77.45 & 76.62 & 0.08\\
		& RANSAC-1M \cite{fischler1981random} & 64.20 & 4.05 & 11.35 & 63.96 & 57.90 & 60.13 & 88.42 & 3.05 & 9.42 & 77.96 & 79.86 & 78.55 & 0.97 \\
		& RANSAC-2M \cite{fischler1981random} & 65.25 & 4.07 & 11.56 & 64.41 & 58.37 & 60.51 & 90.88 & 2.71 & 8.31 & 78.52 & 83.52 & 80.68 & 1.63 \\
		& RANSAC-4M \cite{fischler1981random} & 66.10 & 3.95 & 11.03 & 64.27 & 59.10 & 61.02 & 91.44 & 2.69 & 8.38 & 78.88 & 83.88 & 81.04  & 2.86 \\
		& CG-SAC \cite{quan2020compatibility} & 78.00 & 2.40 & 6.89 & 68.07 & 67.32 & 67.52 & 87.52 & 2.42 & 7.66 & 75.32 & 84.61 & 79.90 & 0.27 \\
		& Ours & \textcolor{Red}{83.98} & 2.18 & \textcolor{Blue}{6.56} & \textcolor{Blue}{72.48} & \textcolor{Red}{78.33} & \textcolor{Red}{75.10} & \textcolor{Red}{93.28} & \textcolor{Red}{2.08} & \textcolor{Blue}{6.55} & 78.94 & \textcolor{Blue}{86.39} & \textcolor{Red}{82.20} & 0.11\\ \hline

	\end{tabular}
	\caption{Registration accuracy results on KITTI dataset compared with other methods}
	\vspace{-3mm}
	\label{tab:eval-table}
\end{table*}

### 2025-04-21 sec/roter-faden.tex
Started convex relaxation based solver (MA thesis) paper

Globally optimal Point cloud registration with intersection graphs.

Intro:

Why ? Fast global optimization of the Truncated least squares function solves the point cloud registration problem, even in case of extreme outlier rates as shown previously. The TLS function is better than maximizing the number of inliers (maximum consensus) since, this has been shown previously many times, e.g. in MAC (CVPR 2024), but also in older works such as MLESAC. The TLS function is easily resilient up to 90\% outliers, a property not achieved by many other problem formulations.

We focus on optimizing the TLS objective and do to deal with the properties of it as a statistical estimator -- it has been shown to perform very well.

We are interested in global algorithms with correctness guarantees, i.e. certifiable algorithms. The algorithm should always output a valid upper bound on the estimation error. There already many heuristic algorithms out there.

All previous algorithms were very slow, we show the first algorithm that runs in real-time, meaning under 100ms on instances that are large enough to be of practical relevance. 
Many methods for global optimization have been  attempted, Branch-and-Bound

The SDP relaxation have led to algorithms dubbed the \textit{first polynomial-time algorithm}. Altought it is true that their worst-case runtime is polynomial due to the fact that they are solvable with interior point solvers, the SDP relaxation is not able to solve all instances. ... (TODO write from intro)

We propose in this paper a geometric approach that enumerates 
convex regions combined with Branch-and-Bound to prune candidates.

This method is more generally known as enumeration of convex regions, first proposed in TODO. It can be implemented easily for the 1D-case, known as the interval stabbing algorithm (TODO cite bustos and CHin and Yang 2020). It was however previously unknown how to generalize it to multiple dimensions, only 1D and 2D versions (TODO yang et al) existed.

We are to our knowledge the first to use concept of an intersection graph to model this geometric problem. It turns out that intersection graphs are very well suited to model outlier-robust TLS problems since they arise naturally from the same geometric problem.

By using maximal clique enumeration on intersection graphs, we can enumerate the convex regions and therefore the local minimas, yielding the global optimum.

Intersection graphs also are a handy tool enabling us to rigorously analyze our algorithm and prove it's global optimality. We also show that our intersection graph is equivalent to the recently proposed compatibility graph that were motivated solely by geometric  observations.

Main contributions: 
- A real-time capable algorithm for solving TLS point cloud registration to certifiable global optimality 
- General approach for enumerating the convex regions of the TLS objective using maximal cliques on intersection graphs 
- Evaluation on adversarial instances for showing global optimality

Related work.

Heuristic methods: Many heuristic methods attepmt to find local only solutions, given. 
Ransac is slow and reuiqres exponential time but recently due to the advances in understanding the geometry of point cloud registration, it has been made robust to over 90\% outliers (cite RANSIC etc.)
BUT Other methods usually considered maximum consensus, that often finds a worse solution that Truncated Least Squares estimator

3. Our stuff:

Problem definition: truncated least squares point cloud registraion. 
How do we solve this ? We first define what inliers are: Given the globally optimal transform, they are all terms that are not truncated. 
Second observation: Knowing the inliers essentially solves the TLS problem since the only thing remaining is to compute the least-squares solution 
using only these inliers. 
Therefore, given the optimal inlier set, we obtain therefore the globap otpimum. 
We can therefore find the global optimum by simply enumerating all inlier sets and solving with least-squares.

- T-sets. We define the transformation sets ...
- The intersection graph
- An inlier set is a non-empty instersection of T-Sets
- Checking for non-empty intersection can be done using least-squares solving
- Inlier sets are cliques in the grap -> 
- The inlier set is maximal

Pruning the search space using BnB

To speed up the clique enumeration, i.e. the enumeration of local optimas, we use branch-and-bound. 
Our use of Branch-and-Bound differs however from many other proposed methods. Recall that lemma 1 states that we can either solve for the (contious) 3D-transformation or the (discrete) inlier set. We have therefore two "perspectives" on the problem: The continous and the discrete.

We do not use BnB for branching over continuous space of 3D-transforms that, but instead branch only over the 3D-space of rotations.

We do not use branching over rotations directly to narrow down the rotational space but instead to prune the intersection graph (i.e. narrow the combinatorial search space). 
At the same time, we use also use the (pruned) intersection graph to compute the degeneracy-based lower bound, pruning additionally the rotational space.

Pruning edges inside rotation nodes.

The

### 2025-04-21 sec/wls_relax.tex
Started convex relaxation based solver (MA thesis) paper

\subsection{Weighted Least Squares (WLS) - Convex Relaxation}
In the following, we derive a convex relaxation for the TLS problem \ref{eq:pcr-tls} over a given interval $\mathcal{X}_r \subseteq \SOthree$. To show the global optimality of Branch-and-Bound, we only need to show that the proposed function is indeed a convex relaxation.

\begin{theorem}
	\label{thm:wls-relax}
	Given are the minimum and maximum of $r_i = ||\mathbf{a}_i - \mathbf{R} \mathbf{b}_i||_2^2$ over all $\mathbf{R} \in \mathcal{X}_r$, named  $r_{min}^i$ and $r_{max}^i$. Then, $\text{\WLSRelax}$ is a convex  relaxation of \ref{eq:pcr-tls} over $\mathcal{X}_r$:
	\begin{equation}
		\label{eq:tls-wls-relaxation}
		\begin{aligned}	
			\mathrm{\WLSRelax}:\quad
			\sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right) + \sum_{i=1}^{N} o_i\epsilon^2\\
		\end{aligned}
	\end{equation}

	with the weights $w_i$ defined as:
	\begin{equation}
		\begin{aligned}	
			w_i  &= 
			\begin{cases}
				\frac{\epsilon^2 - r_{min}^i}{r_{max}^i - r_{min}^i} &  r_{max}^i > \epsilon^2\\
				1 & \, \text{otherwise}
			\end{cases}
		\end{aligned}
	\end{equation}

	and the binary variables $o_i$ deciding whether the i'th residual is certainly an outlier (1) or not (0):

	\begin{equation}
		\begin{aligned}	
			o_i &= \begin{cases}
				1 &  r_{min}^i > \epsilon^2\\
				0 & \, \text{otherwise}
			\end{cases}
		\end{aligned}
	\end{equation}

\end{theorem}

The idea of our relaxation is that non-convexity arises only because of truncation -- we therefore apply an affine transformation to each $r_i$ such that it $\leq \epsilon^2$ over the given interval and thus each term becomes convex. A proof for this relaxation is given in appendix \ref{proof:tls-convex-wls-relaxation}.

The interval $[r_{min}^i, r_{max}^i]$ of the residuals can be computed easily using analytic formulas: Given spherical intervals (balls) with center $\mathbf{R}_c$ and radius $n_r$, for the rotation model $||\mathbf{a}_i - \mathbf{R} \mathbf{b}_i||_2^2$, they are (see appendix \ref{thm:residual-rotation} for a proof):
\begin{equation}
	\label{eq:residual-minmax-rotation}
	\begin{aligned}
		r_{min}^i = \normsq{\mathbf{a}_i} + \normsq{\mathbf{b}_i} - 2 \, ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\max(\theta - n_r, 0))\\
		r_{max}^i = \normsq{\mathbf{a}_i} + \normsq{\mathbf{b}_i} - 2 \, ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\min(\theta + n_r, \pi))
	\end{aligned}
\end{equation}

with 
\begin{equation}
	\begin{aligned}
		\theta = \arccos \left(\frac{(\mathbf{R}_c \, \mathbf{a}_i)^\transposed\mathbf{b}_i}{\norm{\mathbf{a}_i} \, \norm{\mathbf{b}_i}} \right)
	\end{aligned}
\end{equation}

Our convex relaxation can be computed very fast: We only need to compute the interval and weights for each point, which takes linear time in the number of points $N$, i.e. $\mathcal{O}(N)$.

\subsection{Minimizing WLS-Relaxation}
After computing our convex relaxation, we need to find it's global optimum to obtain a lower bound.
Since this is only a weighted least squares problem, it can be solved in practice very fast with a specialized non-iterative solver.\\
\textbf{The lower bound sub-problem}. Since the relaxation is valid only over a given interval (the node of BnB), we need to consider however a inequlity-constrained version of this problem to find the lower bound. 
Still, finding the global optimum of the WLS relaxation remains computationally very easy -- it takes again only linear time and is in practice extremely fast, requiring only 100$\mu s$ - 300$\mu s$.

The feasible set of the problem is a ball $\MB(\vc, n_r)$ with center $\mathbf{c}$ and radius $n_r$, defined over the Euclidean space as $\MB(\vc, n_r) :=\{\mathbf{x} \in \Rone^n : || \mathbf{c} - \mathbf{x}|| \leq n_r \}$. Over non-Euclidean space of rotations, using the geodesic distance $d_{\angle}(\mathbf{R}_c, \mathbf{R})$, we use the similar definition: $\{\mathbf{R} \in \SOthree : d_{\angle}(\mathbf{R}_c, \mathbf{R}) \leq n_r\}$. 
We introduce now the constraint 
$d_{\angle}(\mathbf{R}_c, \mathbf{R}) \leq n_r$, that enforces the solution to be inside the ball where the relaxation is valid.

The problem of minimizing the WLS convex relaxation becomes a a weighted least squares problem (omitting constant terms):

\begin{equation}
	\label{eq:tls-r-opt-branch}
	\begin{aligned}
		\min_{\mathbf{R} \in \SOthree} \quad &\sum_{i=1}^{N} w_i \normsq{\mathbf{a}_i - \mathbf{R}\mathbf{b}_i}\\
		\text{subject to} \quad  &d_{\angle}(\mathbf{R}_c, \mathbf{R}) \leq n_r
	\end{aligned}
\end{equation}
where $\mathbf{R}_c$ is the node center and $n_r$ the node radius.\\
%(Note that $d_{\angle}(\mathbf{R}_c, \mathbf{R}) \leq n_r \iff \trace(\mathbf{R}_c^\transposed \mathbf{R}) \geq 2 \cos(n_r) - 1$)
\textbf{Active-set solver for the rotation sub-problem}. Without the constraint, this problem is the well-known \textit{Wahba-Problem} that is easily solvable with the SVD-algorithm \cite{Kabsch-1976-Point-set-alignment} \cite{8594296} \cite{Lawrence2019APA} \cite{sorkine2017least}.

We combine this observation with the idea of the \textit{active-set} method to solve the constrained problem by splitting it into two simpler sub-problems: First, we just ignore the constraint and solve it with the SVD-algorithm. If the constraint happens to be satisfied, we already have the solution. If not, we know that the inquality-constraint has to be satisfied as an equality \cite[p.467]{Numerical-Optimization-Nocedal-Wright}.
The equality-constraint then requires the rotation to have a rotation angle of exactly $n_r$. 
Since every rotation can be described as a rotation over an axis and angle (Euler's theorem) and the rotation angle is already given by the constraint, the problem simplifies to only finding the axis.\\
\textbf{Solving the equality-constrained case}.
Solving the equality-constrained version of problem \ref{eq:tls-r-opt-branch} is equivalent to the following non-convex problem over the rotation axis $\mathbf{n}$:

\begin{equation}
	\label{eq:rot-est-constrained-qcqp-main}
	\begin{aligned}
		\argmin_{\mathbf{n} \in \Rthree} \quad &-\left(\mathbf{n}^\transposed \mathbf{A} \mathbf{n} + 2 \mathbf{g}^\transposed \mathbf{n} \right)\\
		\text{subject to} \quad &\norm{\mathbf{n}} = 1\\
	\end{aligned}
\end{equation}
A proof of equivalence including the exact definition of $\mathbf{A}$ and $\mathbf{g}$ are given in appendix \ref{proof:wls-relax-eq-constrained-rotation}. This problem is based on the quaternion solution to the Wahba-problem (\textit {Davenport's Q}) \cite{davenportsQ} \cite{8594296} and the axis-angle definition of quaternions for separating out the axis.

The optimization problem \ref{eq:rot-est-constrained-qcqp-main} can again be solved very efficiently by an eigenvalue-decomposition of the 3x3 cross-correlation matrix (that was already computed by the SVD-algorithm) with a subsequent root-finding \cite{10.1007/978-3-642-75536-1_57}. Solving the equality-constrained therefore even takes \textit{constant} time, i.e. is not dependent on the number of points at all.

Overall, our active-set solver for minimizing the convex relaxation is extremely efficient -- in finds the global optimum in linear time.

### 2025-04-21 sec/X_suppl.tex
Added from MA

@@
 \maketitlesupplementary
-
-\begin{theorem}[\textbf{Knowing the ground-truth inliers solves the TLS-problem}]
-	Given the ground-truth inlier set $\mathcal{I}^*$, the global optimum $(\MR^*, \vt^*)$ is found by solving the least-squares problem:
-	\begin{equation}
-		\begin{aligned}
-			(\MR^*, \vt^*) = \argmin_{(\MR, \vt) \in \SEthree} \sum_{i \in \mathcal{I}^*} \normsq{\MR \vp_i - \vq_i + \vt}
-		\end{aligned}
-	\end{equation}
-\end{theorem}
-
-\begin{proof}
-    \label{proof:gt-corrs-solves-tls}
-	Since we know the ground-truth inlier set, we know which residuals are less or equal $\epssq$ and therefore not truncated.
-Since $\min(x, \epssq) = x$ for all $x \leq \epssq$, we can equivalently rewrite the TLS-objective \ref{eq:pcr-tls} as:
-    \begin{equation}
-	\begin{aligned}
-		&\sum_{i=1}^{N} \min \left(\normsq{\MR \vp_i - \vq_i + \vt}, \epsilon^2 \right)\\
-        &= \sum_{i \in \mathcal{I}^*} \normsq{\MR \vp_i - \vq_i  + \vt}  + \sum_{i \in \{1, \cdots, N\}  \setminus \mathcal{I}^*} \epsilon^2
-	\end{aligned}
-    \end{equation}
-    Since the second sum is constant, it does not affect the minimizer of the least-squares term (the first sum).
-Therefore, minimizing only  the first, i.e.
-solving the least-squares problem solves the TLS-problem.
-\end{proof}
 
@@
 
-\begin{lemma}[\textbf{Least-squares solution with two points}]\label{lemma:ls-two-points}
-	In the special case of two points, the least squares solution for point cloud registration is
+\section{Interval analysis of residuals under translation}
+\label{proof:tls-relax-minmax-r}
+
+\begin{theorem}
+	The residual term $r_i = ||\mathbf{a}_i - \mathbf{b}_i + \mathbf{t}||_2^2$
+	has the minimum $r_{min}^i = \max(0, ||\mathbf{d}||_2 - n_r)^2$ and the maximum $r_{max}^i = (||\mathbf{d}||_2 + n_r)^2$
+	for any translation $\mathbf{t} \in \Rthree, ||\mathbf{c} - \mathbf{t}|| \leq n_r$ where $\mathbf{d} = \mathbf{a}_i - \mathbf{b}_i + \mathbf{c}$.
+\end{theorem}
+
+\begin{proof}
+	With $\mathbf{t}= \mathbf{c} - \Delta \mathbf{t} \iff \Delta \mathbf{t} = \mathbf{c} - \mathbf{t}$:
 	\begin{equation}
 		\begin{aligned}
-			\min_{(\MR, \vt) \in \SEthree} & \normsq{\MR \vp_i - \vq_i  + \vt} +  \normsq{\MR \vp_j - \vq_j  + \vt}\\
-			&= \frac{1}{2} \left( \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} \right)^2
+			r_i &= ||\mathbf{a}_i - \mathbf{b}_i + \mathbf{t}||_2^2
+			= ||\mathbf{a}_i - \mathbf{b}_i + \mathbf{c} - \Delta\mathbf{t}||_2^2
+			= ||\mathbf{d} - \Delta\mathbf{t}||_2^2\\ 
+			&= (\mathbf{d} - \Delta\mathbf{t})^\transposed(\mathbf{d} - \Delta\mathbf{t})=
+			\mathbf{d}^\transposed\mathbf{d} - 2 \mathbf{d}^\transposed \Delta\mathbf{t} + \Delta\mathbf{t}^\transposed \Delta\mathbf{t}\\
+			&= ||\mathbf{d}||_2^2 - 2 \, ||\mathbf{d}|| \, ||\Delta\mathbf{t}||  \cos \theta + ||\Delta\mathbf{t}||_2^2\\
 		\end{aligned}
 	\end{equation}
-	Both residual terms are equal at the minimum.
-\end{lemma}
+
+	\begin{equation}
+		\begin{aligned}
+			r_{min}^i = \min_{\Delta \mathbf{t}} r_i &= 
+			||\mathbf{d}||_2^2 + 
+			\min_{\theta \in [-\pi, \pi], ||\Delta\mathbf{t}|| \in [0, n_r]} \left( -2 \, ||\mathbf{d}||  \, ||\Delta\mathbf{t}|| \cos \theta + ||\Delta\mathbf{t}||_2^2 \right) \\
+			&= ||\mathbf{d}||_2^2 + \min_{||\Delta\mathbf{t}|| \in [0, n_r]} \left( -2 \, ||\mathbf{d}||  \, ||\Delta\mathbf{t}|| + ||\Delta\mathbf{t}||_2^2 \right) \\
+			&= \min_{||\Delta\mathbf{t}|| \in [0, n_r]}  (||\mathbf{d}||_2 - ||\Delta\mathbf{t}||_2)^2\\
+			&=\max(0, ||\mathbf{d}||_2 - n_r)^2\\
+			r_{min}^i = \max_{\Delta \mathbf{t}} r_i &= 
+			||\mathbf{d}||_2^2 + 
+			\max_{\theta \in [-\pi, \pi], ||\Delta\mathbf{t}|| \in [0, n_r]} \left( -2 \, ||\mathbf{d}||  \, ||\Delta\mathbf{t}|| \cos \theta + ||\Delta\mathbf{t}||_2^2 \right) \\
+			&= ||\mathbf{d}||_2^2 + 2 \, ||\mathbf{d}||\, n_r + n_r^2 = (||\mathbf{d}||_2 + n_r)^2
+		\end{aligned}
+	\end{equation}
+
+\end{proof}
+
+\section{Interval analysis of residuals under rotation}
+\label{proof:wls-interval-analysis-rotation}
+
+\begin{theorem}
+
+	The residual term $r_i = ||\mathbf{a}_i - \mathbf{R}\mathbf{b}_i ||_2^2$
+	has the minimum $r_{min}^i$ and the maximum $r_{max}^i$
+	over all 3D-rotations $\mathbf{R} \in \SOthree$ that satisfy $d_{\angle}(\mathbf{R}, \mathbf{R_c}) \leq n_r$ where $\mathbf{R_c} \in \SOthree$ is the ball center:
+	\begin{equation}
+		\begin{aligned}
+			r_{min}^i &= ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\max(\theta - n_r, 0))\\
+			r_{max}^i &= ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\min(\theta + n_r, \pi))
+		\end{aligned}
+	\end{equation}
+	with 
+	\begin{equation}
+		\begin{aligned}
+			\theta &= \arccos \left( \frac{(\mathbf{R}_c^\transposed \, \mathbf{a}_i )^\transposed \mathbf{b}_i}
+			{||\mathbf{a}_i|| \, || \mathbf{b}_i||} \right)
+		\end{aligned}
+	\end{equation}
+	being the angle between $\mathbf{R}_c^\transposed \, \mathbf{a}_i $ and $\mathbf{b}_i$.
+
+\end{theorem}
 
 \begin{proof}
-	A well known result is that the optimal translation is related to the optimal rotation and the centroids of both point clouds (see e.g \cite{Kabsch-1976-Point-set-alignment, sorkine2017least}):
+	With $\mathbf{R}  = \mathbf{R}_c \Delta\mathbf{R} \iff \mathbf{R}_c^\transposed\mathbf{R} = \Delta \mathbf{R}$:
 	\begin{equation}
 		\begin{aligned}
-			\vt &= \bar{\vq} - \MR \bar{\vp}\\
-			\bar{\vp} &= \frac{\vp_i + \vp_j}{2}, \, \bar{\vq} = \frac{\vq_i + \vq_j}{2}
+			r_i &= ||\mathbf{a}_i - \mathbf{R}\mathbf{b}_i ||_2^2 \\
+			&= || \mathbf{R}_c^\transposed \left(\mathbf{a}_i - \mathbf{R}\mathbf{b}_i\right) ||_2^2\\
+			&= || \mathbf{R}_c^\transposed \mathbf{a}_i - \mathbf{R}_c^\transposed \mathbf{R}\mathbf{b}_i ||_2^2\\
+			&= \left(\mathbf{R}_c^\transposed \mathbf{a}_i - \Delta \mathbf{R}\mathbf{b}_i \right)^\transposed
+			\left(\mathbf{R}_c^\transposed \mathbf{a}_i - \Delta \mathbf{R}\mathbf{b}_i \right)\\ 
+			&= \left(\mathbf{R}_c^\transposed \mathbf{a}_i \right)^\transposed \left(\mathbf{R}_c^\transposed \mathbf{a}_i \right) - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i + (\Delta \mathbf{R}\mathbf{b}_i )^\transposed (\Delta \mathbf{R}\mathbf{b}_i )\\
+			&= \mathbf{a}_i^\transposed \mathbf{R}_c \mathbf{R}_c^\transposed \mathbf{a}_i - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i + \mathbf{b}_i^\transposed  \Delta \mathbf{R}^\transposed  \Delta \mathbf{R}\mathbf{b}_i\\
+			&= \mathbf{a}_i^\transposed \mathbf{a}_i - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i + \mathbf{b}_i^\transposed \mathbf{b}_i\\
+			&= ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i
 		\end{aligned}
 	\end{equation}
-	We insert the optimal translation into the objective and simplify:
-	\begin{equation}\label{eq:insert-t-opt}
+	Since $\Delta \mathbf{R}$ is defined as rotating a point by at most $n_r$ (definition of geodesic distance $d_{\angle}(\mathbf{R}, \mathbf{R_c}) \leq n_r \iff d_{\angle}(\mathbf{I}, \Delta \mathbf{R} \leq n_r$) from this follows:
+	\begin{equation}
 		\begin{aligned}
-			&\normsq{\MR \vp_i - \vq_i  + (\bar{\vq} - \MR \bar{\vp})} +  \normsq{\MR \vp_j - \vq_j  + (\bar{\vq} - \MR \bar{\vp})}\\
-			=&\normsq{\MR (\vp_i - \bar{\vp}) - (\vq_i -\bar{\vq})} + \normsq{\MR (\vp_j - \bar{\vp}) - (\vq_j -\bar{\vq})} \\
-			=&\normsq{\MR \left(\vp_i - \frac{\vp_i + \vp_j}{2} \right) - \left(\vq_i - \frac{\vq_i + \vq_j}{2} \right)}\\
-			+ &\normsq{\MR \left(\vp_j - \frac{\vp_i + \vp_j}{2} \right) - \left(\vq_j -\frac{\vq_i + \vq_j}{2} \right)} \\
-			=&\normsq{ \frac{1}{2}(\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j))}\\
-			+ &\normsq{ \frac{1}{2}(\MR (\vp_j - \vp_i ) - (\vq_j - \vq_i))}\\
-			=& 2\normsq{ \frac{1}{2}(\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j))}\\
-			=& \frac{1}{2}\normsq{\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j)}\\
+			r_{min}^i &= \min_{\Delta \mathbf{R} \in \SOthree, \, d_{\angle}(\mathbf{I}, \Delta \mathbf{R}) \leq n_r} \left( ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i \right)\\
+			&= ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\max(\theta - n_r, 0)) \\
+			r_{max}^i &=
+			\max_{\Delta \mathbf{R} \in \SOthree, \, d_{\angle}(\mathbf{I}, \Delta \mathbf{R}) \leq n_r} \left( ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i \right)\\
+			&=  ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\min(\theta + n_r, \pi))
 		\end{aligned}
 	\end{equation}
-	We just showed that both residual terms are equal at the optimal translation.	
-	What remains is to optimize over the rotation $\MR$. By lemma \ref{lemma:ls-rot}, the minimum over rotations is:
+
+\end{proof}
+
+\begin{theorem}
+
+	For the truncated least-squares function:
+
+	\begin{equation}
+		\label{eq:tls-t-opt-branch2}
+		\begin{aligned}
+			\text{TLS}(\mathbf{x}) = \sum_{i=1}^{N} \min \left(r(\mathbf{x}, \mathbf{x}_i)^2, \epsilon^2 \right)\\
+		\end{aligned}
+	\end{equation}
+
+	where $\mathbf{x}_i$ are data points and $r_i = r(\cdot)^2$ are residual terms convex in $\mathbf{x}$.
+Given are the minimum and maximum value of each residual term $r_i$ over every $\mathbf{x} \in \mathcal{X}$, denoted as $r_{min}^i$ and $r_{max}^i$ respectively.
+Then, $\text{WLS}_{r}$ is a convex relaxation of \ref{eq:tls-t-opt-branch2} over $\mathcal{X}$:
+	\begin{equation}
+		\label{eq:tls-wls-relaxation}
+		\begin{aligned}	
+			\text{WLS}_{r}(\mathbf{x}) &= 
+			\sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right) + \sum_{i=1}^{N} o_i\epsilon^2\\
+		\end{aligned}
+	\end{equation}
+
+	with the weights $w_i$ defined as:
 	\begin{equation}
 		\begin{aligned}	
-			&\min_{(\MR, \vt) \in \SEthree} \normsq{\MR \vp_i - \vq_i  + \vt} +  \normsq{\MR \vp_j - \vq_j  + \vt}\\
-			&=\min_{\MR \in \SOthree} \frac{1}{2}\normsq{\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j)}\\
-			&= \frac{1}{2} \left( \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} \right)^2
+			w_i  = 
+			\begin{cases}
+				\frac{\epsilon^2 - r_{min}^i}{r_{max}^i - r_{min}^i} &  r_{max}^i > \epsilon^2\\
+				1 & \, \text{otherwise}
+			\end{cases}
 		\end{aligned}
 	\end{equation}
+
+	and the binary variables $o_i$ deciding whether the $i$-th residual is an outlier or not:
+	\begin{equation}
+		\begin{aligned}	
+			o_i &= \begin{cases}
+				1 &  r_{min}^i > \epsilon^2\\
+				0 & \, \text{otherwise}
+			\end{cases}
+		\end{aligned}
+	\end{equation}
+
+\end{theorem}
+
+\begin{proof}
+
+	In the truncated-least-squares function, non-convexity is introduced solely by the truncation since the squared terms are convex.
+	Therefore, we only need to show that the re-weighted sum leads to the residuals not being truncated. 
+	For this, we apply an affine transform on each residual so that $w_i r_i + y_0 = r_{min}^i$ for $r_i = r_{min}^i$ and $w_i r_i + y_0 = \epsilon^2$ for $r_i = r_{max}^i$:
+	Subtracting the first from the second equation yields:
+	\begin{equation}
+		\begin{aligned}	
+			&w_i (r_{max}^i - r_{min}^i) = \epsilon^2 - r_{min}^i\\
+			\iff &w_i  = \frac{\epsilon^2 - r_{min}^i}{r_{max}^i - r_{min}^i}
+		\end{aligned}
+	\end{equation}
+	and solving for $y_0$ yields:
+	\begin{equation}
+		\begin{aligned}	
+			w_i r_{min}^i + y_0 &= r_{min}^i\\
+			\iff y_0 &= r_{min}^i - w_i r_{min}^i\\
+			\iff y_0 &= (1- w_i) r_{min}^i
+		\end{aligned}
+	\end{equation}
+
+	Therefore, $w_k r_k + (1 - w_k) r_{min}^k \leq \epsilon^2$ holds for all $k$ for which $o_k = 0$ and therefore these terms are not truncated and thus convex. The sum
+
+	\begin{equation}
+		\begin{aligned}	
+			\sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right)
+		\end{aligned}
+	\end{equation}
+
+	is therefore convex since sums of convex functions are convex \cite[Ch. 2.3.2, p.36]{Boyd_Vandenberghe_2004}.
+
+	Since $\sum_{i=1}^{N} o_i\epsilon^2$ is constant, the overall term \ref{eq:tls-wls-relaxation} also remains convex.
+Since $r_k \in [r^k_{min}, r^k_{max}]$ and $w_k \leq 1$, $w_k r_k  + (1 - w_k) r_{min}^k \leq r_k$ for all $k$ for which $o_k = 0$  Eq.
+\ref{eq:tls-wls-relaxation} is below the objective \ref{eq:tls-t-opt-branch2}.
+Therefore, \ref{eq:tls-wls-relaxation} is a convex relaxation of \ref{eq:tls-t-opt-branch2} over $\mathcal{X}$.
+
 \end{proof}
 
-\begin{theorem}[\textbf{Non-empty intersection of T-Sets is equivalent to compatible TIMs}]
+\section{Least-squares rotation estimation with equality-constrained angle}
+\label{proof:wls-relax-eq-constrained-rotation}
+
+\begin{theorem}
+
+	The least-squares rotation estimation problem 
 	\begin{equation}
+		\label{eq:tls-rwls-a}
 		\begin{aligned}
-			&\mathcal{T}_i \cap \mathcal{T}_j \neq \emptyset\\
-			\iff &\exists (\MR, \vt) \in \SEthree:\\
-			&\norm{\MR \vp_i - \vq_i + \vt} \leq \epsilon\\
-			\land &\norm{\MR \vp_j - \vq_j + \vt} \leq \epsilon\\
-			\iff &| \norm{\va_{ij}} - \norm{\vb_{ij}} | \leq 2 \epsilon
+			\argmin_{\mathbf{R} \in \SOthree}  \quad &\sum_{i=1}^{N} w_i ||\mathbf{R}_c^\transposed \mathbf{a}_i - \Delta\mathbf{R} \mathbf{b}_i ||_2^2 \\
+			\text{subject to} \quad &d_{\angle}(\mathbf{I}, \Delta\mathbf{R}) = n_r
 		\end{aligned}
 	\end{equation}
-	where 
+	where the angle of rotation is constrained to $n_r$ is equivalent to the following optimization problem over the rotation axis $\mathbf{n}$:
 	\begin{equation}
-		\label{eq:tims}
+		\label{eq:rot-est-constrained-qcqp}
 		\begin{aligned}
-			\va_{ij} &= \vp_i - \vp_j\\
-			\vb_{ij} &= \vq_i - \vq_j
+			\argmin_{\mathbf{n} \in \Rthree} \quad &-\left(\mathbf{n}^\transposed \mathbf{A} \mathbf{n} + 2 \mathbf{g}^\transposed \mathbf{n} \right)\\
+			\text{subject to} \quad &|| \mathbf{n}|| = 1\\
+		\end{aligned}
+	\end{equation}
+	with
+	\begin{equation}
+		\label{eq:ev-constr-data-mat-a}
+		\begin{aligned}
+			\mathbf{B} &= \sum_{i=1}^N w_i \mathbf{a}_i\mathbf{b}_i^\transposed \in \Rone^{3 \times 3}, \quad
+			\mathbf{z} = \sum_{i=1}^N w_i [\mathbf{a}_i]_{\times}\mathbf{b}_i \in \Rone^{3}, \quad \mathbf{C} = \mathbf{B} + \mathbf{B}^\transposed\\
+			\mathbf{A} &= \sin^2 \left(\frac{n_r}{2} \right) \mathbf{C}, \quad \mathbf{g} =  \sin \left(\frac{n_r}{2} \right) \cos \left(\frac{n_r}{2} \right) \mathbf{z}
 		\end{aligned}
@@
 \begin{proof}
-	\label{proof:non-empty-intersection-implies-tims-compat}
-	Whether there exists a $(\MR, \vt) \in \SEthree$ such that $\norm{\MR \vp_i - \vq_i  + \vt} \leq \epsilon$ is equivalent to whether the minimum is $\leq \epsilon$, i.e.
-whether $\min_{(\MR, \vt) \in \SEthree} \norm{\MR \vp_i - \vq_i  + \vt} \leq \epsilon$ holds.
-	By lemma \ref{lemma:ls-two-points} by minimizing the sum of two residuals, at the minimum they are both equal and therefore if their sum is $\leq 2\epssq$, they both are $\leq \epssq$.
-We conclude therefore
+
+	We first separate the optimization over the rotation axis and angle.
+For this, we use the quaternion representation of \ref{eq:tls-rwls-a}.
+One can show that the problem \ref{eq:tls-rwls-a} is equivalent to:
+
 	\begin{equation}
+		\label{eq:davenports-q-problem}
 		\begin{aligned}
-			&\mathcal{T}_i \cap \mathcal{T}_j \neq \emptyset \iff\\
-			&\min_{(\MR, \vt) \in \SEthree} \normsq{\MR \vp_i - \vq_i  + \vt} + \normsq{\MR \vp_j - \vq_j  + \vt} \leq 2\epssq\\
-			& \iff \frac{1}{2} \left( \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} \right)^2 \leq 2\epssq\\
-			& \iff | \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} | \leq 2\epsilon\\
-			& \iff | \norm{\va_{ij}} - \norm{\vb_{ij}} | \leq 2 \epsilon
+			\max_{\mathbf{q} \in \Rone^4} &\quad \mathbf{q}^\transposed \mathbf{K} \mathbf{q}\\
+			\text{subject to} &\quad  ||\mathbf{q}|| = 1
 		\end{aligned}
 	\end{equation}
-\end{proof}
 
-\begin{lemma}[\textbf{Consensus sets are cliques in the intersection graph}]
-	Consensus sets are cliques in the intersection graph.
-\end{lemma}
+	where $\mathbf{q}$ is a 4-dimensional vector representing the quaternion of rotation and $\mathbf{K} \in \Rone^{4\times4}$ is a data matrix.
+This optimization problem is known as the \textit{Davenport's q-method}.
+The quaternion can be written in axis-angle form as:
 
-\begin{proof}
-	Cliques in the intersection imply by definition pairwise non-empty intersection of T-Sets.
-Since pairwise non-empty intersection is a necessary condition for total non-empty intersection and therefore consensus sets, consensus sets are cliques in the intersection graph.
-\end{proof}
-
-\begin{lemma}[\textbf{Maximal clique enumeration and global optimality}]
-	\label{lemma:max-clique-non-helly-app}
-	Maximal clique enumeration finds the ground-truth consensus set and therefore the global minimum of an instance if all maximal cliques of this instance are maximal consensus sets.
-\end{lemma}
-
-\begin{proof}
-	The key statement of the lemma is that if every maximal clique is a maximal consensus set, then subcliques of maximal cliques (which by definition are not maximal cliques) cannot be the ground truth inlier set.
-Consequently, they need not be enumerated.
-
-	The global optimum lies in the union of all T-sets, i.e.
-$\bigcup_{i \in \RangeOneToN} \mathcal{T}_i$.
-Every vertex of a graph is contained in at least one maximal clique of that graph.
-Therefore, the global optimum is contained in at least one union $\bigcup_{i \in \vc} \mathcal{T}_i$ of a maximal clique $\vc$.
-
-	The intersection of the T-sets of any subclique contains the intersection of the T-sets of a maximal clique containing that subclique.
-Therefore, if the consensus set is maximal, the best solution (minimizing the TLS objective) that can be found among subcliques is no better than this maximal clique.
-If this holds for every maximal clique of an instance, then the ground-truth inlier set is not a subclique of any maximal clique, and therefore must itself be a maximal clique.
-Thus, enumerating maximal cliques will find the ground-truth consensus set if all maximal cliques of this instance are maximal consensus sets.
-v
-\end{proof}
-
-\label{proof:wls-interval-analysis-rotation}
-
-\begin{lemma}[\textbf{Consensus set size and TLS objective }]
-	Given an upper-bound $\text{UB}$ on the TLS objective, the ground-truth consensus set must have at least $d_{m}$:
 	\begin{equation}
+		\label{eq:q-axis-angle}
 		\begin{aligned}
-            d_m = \lceil N - \text{UB} / \epssq \rceil
+			\mathbf{q} = \left( \sin \left( \frac{\theta}{2} \right) \mathbf{n}, \cos \left(\frac{\theta}{2} \right) \right)^\transposed
 		\end{aligned}
 	\end{equation}
-\end{lemma}
+	i.e.
+the first three elements are the unit-vector $\mathbf{n} \in \Rthree, ||\mathbf{n}|| = 1$ of the axis of rotation multiplied by $\sin \left( \frac{\theta}{2} \right)$, while the fourth element is just $\cos \left(\frac{\theta}{2} \right)$ and $\theta$ is the angle of rotation.
+	The data matrix $\mathbf{K}$ has the following block-structure:
 
-\begin{proof}
-    By the definition of the TLS objective, 
-    $r_{in} + (N - d_m) \epsilon^2 \leq \text{UB}$ must hold, where $r_{in}$ is the (unknown) sum of the inlier residuals. Rearranging yields $d_m \geq r_{in}  + N - \text{UB} / \epssq$.
-    Since all the summands in the TLS objective are non-negative, $r_{in} = 0$ is a valid lower bound.
-Since the number of elements in the consensus set $d_m$ must be an integer, we conclude that $d_m = \lceil N - \text{UB} / \epssq \rceil$.
+	\begin{equation}
+		\label{eq:k-mat}
+		\begin{aligned}
+			\mathbf{K} = \left(
+			\begin{array}{c|c}
+				\begin{array}{ccc}
+					& & \\
+					& \mathbf{C} & \\
+					& & 
+				\end{array}
+				&
+				\begin{array}{c}
+					\mathbf{z} \\
+				\end{array}
+				\\
+				\hline
+				\begin{array}{ccc}
+					& \mathbf{z}^\transposed &
+				\end{array}
+				&
+				\trace(\mathbf{C})
+			\end{array}
+			\right) \in \Rone^{4 \times 4}
+		\end{aligned}
+	\end{equation}
+
+	We can rewrite the objective function $\mathbf{q}^\transposed \mathbf{K} \mathbf{q}$  of \ref{eq:davenports-q-problem} using the blocks of the block-matrix $\mathbf{K}$:
+
+	\begin{equation}
+		\label{eq:dv-q-rew}
+		\begin{aligned}
+			&\mathbf{q}^\transposed \mathbf{K} \mathbf{q}\\
+			&= \sin^2 \left(\frac{\theta}{2} \right) \mathbf{n}^\transposed \mathbf{C} \mathbf{n} + 2 \sin \left(\frac{\theta}{2} \right) \cos \left(\frac{\theta}{2} \right) \mathbf{z}^\transposed \mathbf{n} + \cos^2 \left(\frac{\theta}{2} \right) \trace(\mathbf{C})
+		\end{aligned}
+	\end{equation}
+
+	We now have a scalar term where the axis $\mathbf{n}$ and the angle $\theta$ occur separately and with this we can formulate the optimization problem over the axis.
+By setting $\theta = n_r$ and solving only over $\mathbf{n}$ we can therefore solve the equality-constrained problem \ref{eq:tls-rwls-a}.
+
+	Since the term $ \cos^2 \left(\frac{\theta}{2} \right) \trace(\mathbf{C})$ is constant in Eq. \ref{eq:dv-q-rew}, it does no affect the minimizer and we will leave it out. 
+	By defining the data-matrices as in \ref{eq:ev-constr-data-mat-a}, we immediatelly see that the problem 
+	\ref{eq:rot-est-constrained-qcqp} is equivalent to Davenports Q-problem \ref{eq:davenports-q-problem} and therefore equivalent to \ref{eq:tls-rwls-a}, concluding the proof.
 \end{proof}

### 2025-04-21 sec/2_related_work.tex
Structure

@@
 Exact methods that guarantee to find the global optimum are mostly based on Branch-and-Bound (BnB) or semidefinite relaxations. A common formulation, closely related to the TLS objective, is the
-Maximum Consensus (MC), one where the number of inliers is maximized.
+Maximum Consensus (MC) one, where the number of inliers is maximized.
 Branch-and-Bound has been proposed for rotation-only solving of this problem \cite{4408896, 10.1007/978-3-642-37444-9_42}.
@@
 \cite{9785843} propose a sparse SPD relaxation for (among other problems) the full 6-DoF TLS point cloud registration problem and also a fast rank-1 SDP solver \textit{STRIDE}.
-They can solve synthetic instances with $N=100$ points to global optimality with up to 80\% outlier rate, but require over 200 seconds.
+They can solve synthetic instances with $N=100$ points to global optimality with up to 80\% outlier rate, but this requires 200 seconds.
 SDPs have been used for similar problems such as multiple point cloud registration \cite{9157383}, but with similarly large runtimes.

### 2025-04-21 sec/3_tls_opt.tex
Structure

@@
 \subsection{The Truncated Least Squares problem}
-The Truncated Least Squares (TLS) problem is a least squares esitmation problem where residuals are truncadet above a threshold.
-(It is similar to the Maximum Consensus problem but additionally considers the residual values, obtaining slightly better estimates.)
-Outliers are handled by truncating residuals above a threshold $\epsilon^2$, yielding the \textit{Truncated Least Squares} (TLS) problem:
-
 \begin{figure}[!ht]
@@
 	\end{adjustbox}
-	\caption{The non-convex Truncated least-squares (TLS) Funktion with multiple local minima.Shown is the function  $\sum_{i=1}^{N}\min((x - y_i)^2, \epsilon^2)$ for $\epsilon^2=0.1$, and nine $y_i$'s (gray lines).}
+	\caption{The non-convex Truncated least-squares (TLS) function with multiple local minima.Shown is the function  $\sum_{i=1}^{N}\min((x - y_i)^2, \epsilon^2)$ for $\epsilon^2=0.1$, and nine $y_i$'s (gray lines).}
 	\label{fig:tlscostmulterm}
@@
 
-Die truncated least squares Funktion ist eine Funktion der Form:
+The Truncated Least Squares (TLS) problem is a least squares estimation problem where residuals are truncated above a threshold.
+(It is similar to the Maximum Consensus problem but additionally considers the residual values, obtaining slightly better estimates.)
+Outliers are handled by truncating residuals above a threshold $\epsilon^2$, the objective function is therefore:
 
@@
 For a proof of this WLS-relaxation, see \ref{proof:tls-convex-wls-relaxation}.
+
+\subsection{Interval analysis of residuals}
+
+In the following we show how $r_{min}$ and $r_{max}$ can easily be computed analytically for different models (i.e. translation and rotation), given spherical intervals, i.e. balls.
+
+The minimum $r_{min}^i$ and maximum $r_{max}^i$ of the residual $r_i =||\mathbf{a}_i - \mathbf{b}_i + \mathbf{t}||_2^2$ (translation) over a ball with center $\mathbf{c}$ and radius $n_r$ are: (see \ref{proof:tls-relax-minmax-r} for a proof):
+
+\begin{equation}
+	\begin{aligned}	
+		r_{min}^i &= \max(0, ||\mathbf{d}_i||_2 - n_r)^2\\
+		r_{max}^i &= (||\mathbf{d}_i||_2 + n_r)^2
+	\end{aligned}
+\end{equation}
+
+with $\mathbf{d}_i$ defined as:
+
+\begin{equation}
+	\begin{aligned}	
+		\mathbf{d}_i = \mathbf{a}_i - \mathbf{b}_i + \mathbf{c}
+	\end{aligned}
+\end{equation}
+
+Similarly, for the rotation residual $||\mathbf{a}_i - \mathbf{R}\mathbf{b}_i ||_2^2$ over a over a ball with center $\mathbf{R}_c$ and (geodesic) radius $n_r$ they are (see \ref{proof:wls-interval-analysis-rotation} for a proof):
+
+\begin{equation}
+	\label{eq:residual-minmax-rotation}
+	\begin{aligned}
+		r_{min}^i = ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\max(\theta - n_r, 0))\\
+		r_{max}^i = ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\min(\theta + n_r, \pi))
+	\end{aligned}
+\end{equation}
+
+where $\theta$ is the angle between $\mathbf{R}_c^\transposed \,\mathbf{a}_i$ and $\mathbf{b}_i$:
+
+\begin{equation}
+	\begin{aligned}
+		\theta = \arccos \left( \frac{(\mathbf{R}_c^\transposed \, \mathbf{a}_i )^\transposed \mathbf{b}_i}{|| \mathbf{a}_i || \, || \mathbf{b}_i ||} \right)
+	\end{aligned}
+\end{equation}
+
+Computing these intervals requires constant time and is in practice very fast when using trigonometric identities to elimitade the the need for evaluating the cosine function.
+
+\section{Minimizing the WLS convex relaxation}
+After having computed the WLS relaxation (i.e. the weights), we need to minimize it to obtain a lower bound on the TLS objective. 
+Since the relaxation is a simple weighted least squares function, minimizing it is rather easy.
+The only thing we need to consider is that we seek the minimum that is inside the Branch and Bound node.
+We therefore need to introduce an inequality constraint into the problem.
+This constrained problem can be solved can still be solved very efficiently by utilizing custom active-set solvers that are only slightly more expensive than solving the least-squares problem.
+
+\subsection{Minimizing the WLS-relaxation over translations}
+
+To minimize the WLS-relaxation over the translation $\mathbf{t}$ for obtaining the lower bound over the ball with center $\mathbf{c}$ and radius $n_r$, we need to solve the following optimization problem:
+\begin{equation}
+	\label{eq:tls-twls-constrained}
+	\begin{aligned}
+		\argmin_{\mathbf{x} \in \Rone^3} \quad &\sum_{i=1}^{N} w_i ||\mathbf{d}_i - \mathbf{x}||_2^2 \\
+		\text{subject to} \quad  &||\mathbf{x}||_2 \leq n_r
+	\end{aligned}
+\end{equation}
+
+where 
+\begin{equation}
+	\begin{aligned}
+		\mathbf{x} = \mathbf{c} - \mathbf{t}  \iff  \mathbf{t} = \mathbf{c} - \mathbf{x}\\
+		\mathbf{d}_i = \mathbf{a}_i - \mathbf{b}_i + \mathbf{c}
+	\end{aligned}
+\end{equation}
+
+The constraint $||\mathbf{x}||_2 \leq n_r$ ensures that the found $\mathbf{x}$ is inside the BnB node.
+The other sums of the convex relaxation \ref{eq:tls-wls-relaxation1} disappeared because they are constant (i.e.
+they do not depend on $\mathbf{x}$).
+This problem is a convex QCQP and therefore solvable by any standard QCQP solver.
+However, we show that the solution can be obtained even more easily in closed form, which in practice results in a much faster algorithm.
+
+We first form the Lagrangian to account for the constraint (which is squared):
+
+\begin{equation}
+	\begin{aligned}	
+		L(\mathbf{x}, \lambda) =  \sum_{i=1}^{N} w_i ||\mathbf{d}_i - \mathbf{x}||_2^2 + \lambda (
+		\mathbf{x}^\transposed \mathbf{x} - n_r^2)\\
+	\end{aligned}
+\end{equation}
+
+then, we set the first partial derivatives to zero: 
+\begin{equation}
+	\begin{aligned}	
+		\frac{\partial L}{\partial \mathbf{x}} &= \frac{\partial }{\partial \mathbf{x}} \left( \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)^\transposed \left(\mathbf{d}_i - \mathbf{x}\right) + \lambda (
+		\mathbf{x}^\transposed \mathbf{x} - n_r^2) \right)\\
+		&= - 2\sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x}\\
+		&= \mathbf{0}\\
+	\end{aligned}
+\end{equation}
+
+By rearranging, we can obtain therefore the solution in closed form:
+\begin{equation}
+	\label{eq:trans-qcqp-lagrangian}
+	\begin{aligned}	
+		-2  &\sum_{i=1}^{N}  w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x} = \mathbf{0}\\
+		\iff \lambda \mathbf{x} &= \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)\\
+		\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \sum_{i=1}^{N}  w_i \mathbf{x}\\
+		\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \mathbf{x} \sum_{i=1}^{N}  w_i\\
+		\left( \lambda + \sum_{i=1}^{N}  w_i \right) \mathbf{x} &= \sum_{i=1}^{N} w_i \mathbf{d}_i \\
+		\mathbf{x} &= \frac{1}{\lambda + \sum_{i=1}^{N}  w_i} \sum_{i=1}^{N} w_i \mathbf{d}_i 
+	\end{aligned}
+\end{equation}
+
+Now the Karush-Kuhn-Tucker (KKT) conditions require $\lambda \geq 0$ and $\lambda (
+\mathbf{x}^\transposed \mathbf{x} - n_r^2) = 0$ in addition to satisfying the constraint, i.e.
+$|| \mathbf{x}||_2 \leq n_r$.
+It follows that in practice we obtain the optimum by first computing $\mathbf{x}$ with Eq.
+\ref{eq:trans-qcqp-lagrangian}, assuming $\lambda = 0$.
+Then we check whether the inequality constraint is satisfied, i.e.
+whether $|| \mathbf{x}||_2 \leq n_r$ holds, and if not, we simply normalise the vector so that $|| \mathbf{x}||_2 = n_r$.
+This gives us the optimal solution vector.
+Intuitively, from eq.
+\ref{eq:trans-qcqp-lagrangian} we already see that the direction of the vector does not change regardless of the optimal Lagrange multiplier $\lambda$ since it is only a scalar factor.
+
+This optimization problem can be understood as a special case of the Trust Region Problem for which strong duality holds \cite[Ch.
+5.2, p.229]{Boyd_Vandenberghe_2004} as well as a regularized linear regression problem.
+
+\subsection{Minimizing the WLS-relaxation over rotations}
+
+To minimize the convex relaxation over rotations and obtain a lower bound for given BnB-node that is a ball with center $\mathbf{R_c}$ and (geodesic) radius $n_r$, we need to solve the following optimization problem:
+
+\begin{equation}
+	\label{eq:tls-rwls}
+	\begin{aligned}
+		\argmin_{\Delta \mathbf{R} \in \SOthree}  \quad &\sum_{i=1}^{N} w_i ||\mathbf{R}_c^\transposed \mathbf{a}_i - \Delta\mathbf{R} \mathbf{b}_i ||_2^2 \\
+		\text{subject to} \quad &d_{\angle}(\mathbf{I}, \Delta\mathbf{R}) \leq n_r
+	\end{aligned}
+\end{equation}
+
+Where $\Delta\mathbf{R}$ is the rotation relative to the node's center:
+
+\begin{equation}
+	\begin{aligned}
+		\Delta\mathbf{R} = \mathbf{R_c}^\transposed\mathbf{R} \iff \mathbf{R_c}\Delta\mathbf{R} = \mathbf{R}
+	\end{aligned}
+\end{equation}
+
+The unconstrained version of this problem is known as the \textit{Wahba-problem} for which a simple SVD-algorithm exists, known as the Kabsch-Umeyama algorithm \cite{Kabsch-1978-Point-set-alignment} \cite{Least-squares-estimation-point-sets-Umeyama-1991} \cite{Lawrence2019APA}.
+But this problem is modified and has in addition an inequality-constraint.
+In the following, we develop a custom solver that solves this problem very efficiently.
+
+The first idea is to just try to solve the unconstrained version and check whether it satisfies the inequality constraint (i.e. the solution is inside the ball). 
+We can do that, but unfortunately the solution does not always satisfy the inequality constraint.
+We gain however an important insight: If the solution satisfies the constraint and lies therefore in the spherical region, we have already found the optimum of the constrained problem.
+If however the optimum of the unconstrained problem is outside of the region, the optimal solution to the constrained problem has to lie on the border of the region.
+It then has to satisfy the constraint as an equality, i.e.
+$ d_{\angle}(\mathbf{I}, \Delta\mathbf{R})  = n_r$ \cite[Ch.15 p.424-427]{Numerical-Optimization-Nocedal-Wright}.
+This two-stage method for solving inequality-constrained nonlinear problems is known as the \textit{active-set} method \cite[Ch.
+16.5, p.467]{Numerical-Optimization-Nocedal-Wright}, an inequality constraint is set to be active if the solution satisfies it as an equality.
+
+Using this active-set approach, we can effectively solve this inequality-constrained optimization problem, since we already have the SVD-algorithm for the unconstrained case.
+And for equality-constrained case, we make the observation that by Euler's theorem that says that every 3D-rotation can be separated into a n axis and angle, we only have to optimize over the axis when the angle is equality-constrained and therefore fixed.
+
+As an algorithm outline, we therefore first solve the unconstrained problem with the SVD-algorithm and check whether it satisfies the inequality-constraint.
+If not, we solve for the equality-constrained problem where we seek for the rotation that has the rotation angle $n_r$, that is more difficult case.
 

### 2025-04-21 sec/X_suppl.tex
Structure

@@
 	has the minimum $r_{min}^i = \max(0, ||\mathbf{d}||_2 - n_r)^2$ and the maximum $r_{max}^i = (||\mathbf{d}||_2 + n_r)^2$
-	for any translation $\mathbf{t} \in \Rthree, ||\mathbf{c} - \mathbf{t}|| \leq n_r$ where $\mathbf{d} = \mathbf{a}_i - \mathbf{b}_i + \mathbf{c}$.
+	for any translation $\mathbf{t} \in \Rthree$ and $||\mathbf{c} - \mathbf{t}|| \leq n_r$ where $\mathbf{d} = \mathbf{a}_i - \mathbf{b}_i + \mathbf{c}$.
 \end{theorem}
@@
 
-	where $\mathbf{x}_i$ are data points and $r_i = r(\cdot)^2$ are residual terms convex in $\mathbf{x}$.
+	where $\mathbf{x}_i$ are data points and $r(\mathbf{x}, \mathbf{x}_i)^2 = r_i$ are residual terms convex in $\mathbf{x}$.
 Given are the minimum and maximum value of each residual term $r_i$ over every $\mathbf{x} \in \mathcal{X}$, denoted as $r_{min}^i$ and $r_{max}^i$ respectively.

### 2025-04-21 sec/3_tls_opt.tex
TODO, eval

@@
 Then we check whether the inequality constraint is satisfied, i.e.
-whether $|| \mathbf{x}||_2 \leq n_r$ holds, and if not, we simply normalise the vector so that $|| \mathbf{x}||_2 = n_r$.
-This gives us the optimal solution vector.
-Intuitively, from eq.
+whether $|| \mathbf{x}||_2 \leq n_r$ holds, and if not, we simply normalise $\mathbf{x}$ so that $|| \mathbf{x}||_2 = n_r$, yielding the optimal solution.
+Intuitively, from Eq.
 \ref{eq:trans-qcqp-lagrangian} we already see that the direction of the vector does not change regardless of the optimal Lagrange multiplier $\lambda$ since it is only a scalar factor.

### 2025-04-21 sec/4_evaluation.tex
TODO, eval

@@
 \ref{tab:eval-kitti}, our method achieves comparable results to other state-of-the-art methods without dataset-specific tuning of parameters, outperformed only by MAC and SC$^2$-PCR.
+
+\section{Experimentelle Evaluation}
+
+In diesem Abschnitt wird die vorgeschlagene konvexe Relaxierung evaluiert hinsichtlich ihres Rechenaufwands und der Qualität.
+Wir messen die Qualität anhand der Anzahl der Iterationen, die Branch-and-Bound (BnB) benötigt, um zu terminieren.
+Zu diesem Zweck haben wir einen einfache Variante von best-first-BnB implementiert, so wie in \cite{10.1007/978-3-642-37444-9_42}.
+
+Wir haben die Methode in C++ unter Verwendung der \textit{Eigen}\footnote{https://eigen.tuxfamily.org/} Matrix-Bibliothek implementiert.
+Wir haben die Implementierung so optimiert, dass die Auswertung trigonometrischer Funktionen durch die Verwendung trigonometrischer Identitäten entfällt.
+Auf diese Weise erfordert die Berechnung der WLS-Relaxierung nur elementare Berechnungen und eine Quadratwurzel-Berechnung pro Punkt.
+Unsere Implementierung verwendet keine Parallelisierung.
+Wir messen die Gesamtzeit für BnB und definieren die $y$-Suboptimalität $y_{subopt}$ als die Suboptimalität des Wertes der Zielfunktionen basierend auf der aktuellen unteren (LB) und oberen Schranke (UB) von BnB \cite{Misener2013}:
+
+\begin{align}
+	y_{subopt} = \frac{\text{UB} - \text{LB}}{|\text{LB}|}
+\end{align}
+
+Wir setzen die Bedingung für Terminierung von BnB bei $y_{subopt} < 10^{-4}$ so wie in \cite{Misener2013}.
+Wir haben die Experimente auf einem PC mit Intel i7-12700K-CPU und 32GiB~RAM durchgeführt.   
+\subsection{Synthetische Experimente}
+
+Mittels synthetischen Experimenten können wir leicht die Robustheit gegenüber Ausreißern testen. 
+Wir haben Instanzen für das Optimierungsproblem Gl.
+\ref{eq:pcr-tls} erzeugt, indem wir zuerst Punkte für die korrekten Messungen aus einer Gleichverteilung gesampled haben wie in \cite{Yang20tro-teaser}.
+Diese korrekten Punkte werden danach skaliert um den Faktor $30$ um ein realistisches Verhältnis aus Ausdehnung (Größe) der Punktwolke und Messgenauigkeit zu erhalten.
+30m ist dabei der kleinste zu erwartende Abstand, in dem noch Punkte von einem LiDAR-sensor detektiert werden.
+Üblich haben LiDAR-Sensoren eine Reichweite von über 100m.
+Als nächstes wird das Messrauschen simuliert, indem aus einer Gleichverteilung im Intervall der Rauschgrenze $\epsilon$ gesampled wird und diese Werte zu jedem Punkt addiert werden.
+Zuletzt werden Ausreißer-Punkte simuliert, indem von einer Gleichverteilung innerhalb der bounding box der korrekten Punkte gesampled wird.
+
+\subsection{Ergebnisse}
+
+\begin{figure}[!ht]
+	\centering
+	\begin{adjustbox}{width=1.\linewidth}
+		\includegraphics{figures/robust_wahba_runtime_rotation_error.pdf}
+		\end{adjustbox}
+		\caption{Der durchschnittliche Rotationsfehler in Abhängigkeit vom Anteil der Ausreißer: Unsere Methode bleibt robust bis zu 95\% Ausreißer, wie von der Truncated Least Squares Formulierung zu erwarten ist.}
+		\label{fig:rotation-error-over-outlier-rate}
+\end{figure}
+
+\begin{figure}[!ht]
+	\centering
+	\begin{adjustbox}{width=1.\linewidth}
+		\includegraphics{figures/robust_wahba_runtime_execution_time.pdf}
+	\end{adjustbox}
+	\caption{Die durchschnittliche Ausführungszeit von Branch-and-Bound in Abhängigkeit von der Gesamtzahl der Punkte $N$ des Optimierungsproblems.}
+	\label{fig:runtime}
+\end{figure}
+
+Wir haben eine Ausreißer-Rate von bis zu $95\%$ untersucht.
+Für jede evaluierte Rate und Datengröße wurden 100 Stichproben generiert.
+Die Ergebnisse des Rotationsfehlers in Abhängigkeit von der Ausreißer-Rate sind in Abb.
+\ref{fig:rotation-error-over-outlier-rate} dargestellt.
+Die Schätzung bleibt korrekt mit einem geringen Fehler von unter $1$ Grad.
+Wir haben auch die Skalierbarkeit unserer Methode über die Gesamtzahl der Punkte $N$ untersucht.
+Wie die Ergebnisse in Abb.
+\ref{fig:runtime} zeigen, kann unser Ansatz mit einer Laufzeit von unter 100 Millisekunden sogar für Echtzeitanwendungen eingesetzt werden.

### 2025-04-21 sec/3_tls_opt.tex
Shorten ineq solver

@@
 
-We can show that this optimization problem is equivalent to the following problem that finds the optimal axis of rotation $\mathbf{n}^*$:
+We show that this optimization problem is equivalent to the following problem that finds the optimal axis of rotation $\mathbf{n}^*$:
 
@@
 
-with the data matrices $\mathbf{A}$ and $\mathbf{g}$ that depend on the points $\mathbf{a}_i$ and $\mathbf{b}_i$ and the angle constraint $n_r$:
+with the data matrices $\mathbf{A}$ and $\mathbf{g}$ that depend on the points $\mathbf{a}_i$ and $\mathbf{b}_i$ and the angle constraint $n_r$.
+A definition is given in the Appendix \ref{thm:constrained-wahba-davenport}.
 
-\begin{equation}
-	\label{eq:davenport-data-matrices}
-	\begin{aligned}
-		\mathbf{B} &= \sum_{i=1}^N w_i \mathbf{R}_c^\transposed \mathbf{a}_i\mathbf{b}_i^\transposed \in \Rone^{3 \times 3}, \quad
-		\mathbf{z} = \sum_{i=1}^N w_i [\mathbf{a}_i]_{\times}\mathbf{b}_i \in \Rone^{3}, \quad \mathbf{C} = \mathbf{B} + \mathbf{B}^\transposed\\
-		\mathbf{A} &= \sin^2 \left(\frac{n_r}{2} \right) \mathbf{C}, \quad \mathbf{g} =  \sin \left(\frac{n_r}{2} \right) \cos \left(\frac{n_r}{2} \right) \mathbf{z}
-	\end{aligned}
-\end{equation}
+To solve this non-convex optimization problem \ref{eq:eq-constrained-prob-ev} over the sphere, we use the algorithm proposed in \cite{10.1007/978-3-642-75536-1_57}.
+We perform overall two steps: We first need to compute the eigenvalue decomposition  $\mathbf{A} = \mathbf{Q}\mathbf{\Lambda}\mathbf{Q}^\transposed$ and then need to find the rightmost root of a certain rational function (characteristic polynomial).
+This can be done in practice very fast, yielding a constant time algorithm.
 
-This derivation is based on the Daveport's Quaternion-method \cite{davenportsQ} \cite{8594296}, see \ref{proof:wls-relax-eq-constrained-rotation} for a proof of the equivalence of both optimization problems.
-
-To solve the optimization problem \ref{eq:eq-constrained-prob-ev}, we perform overall two steps: We first need to compute the eigenvalue decomposition  $\mathbf{A} = \mathbf{Q}\mathbf{\Lambda}\mathbf{Q}^\transposed$ and then need to find largest root of the following rational function $C(\gamma)$ (characteristic polynomial):
-
-\begin{equation}
-	\label{eq:characterisitc-poly}
-	\begin{aligned}
-		C(\gamma) = \sum_{i=1}^3 \left(\frac{a_i}{\lambda_i - \gamma}\right)^2 - 1
-	\end{aligned}
-\end{equation}
-
-with
-\begin{equation}
-	\begin{aligned}
-		\mathbf{a} = (a_1, a_2, a_3)^\transposed = \mathbf{Q}^\transposed \mathbf{g} \in \Rthree, \quad \mathbf{\Lambda} = \text{diag}(\lambda_1, ..., \lambda_3)
-	\end{aligned}
-\end{equation}
-
-\begin{algorithm}[!ht]
-	\caption{Solve constrained eigenvalue problem}
-	\label{alg:solve-constrained-ev}
-	\begin{algorithmic}[1]
-		\Require{Data matrices $\mathbf{A} \in \Rone^{3\times3}, \mathbf{g} \in \Rthree$}
-		\Ensure Solution $\mathbf{n}^*$ to the optimization problem Eq. \ref{eq:eq-constrained-prob-ev}
-		\hypertarget{ref:min-constrained-ev}{
-			\Function{\code{solve\_constrained\_ev}}{$\mathbf{A}, \mathbf{g}$}}
-		\State $\code{kEps} \leftarrow 2^{-48}$ \algorithmiccomment{Epsilon-value for double precision}
-		\State $\mathbf{Q}, \mathbf{\Lambda} \leftarrow \code{eig(} \mathbf{A} \code{)}$ \algorithmiccomment{Compute eigen-decomposition}
-		\State $k = \argmax\{\lambda_1, \lambda_2, \lambda_3\}$
-		\State $\mathbf{a} \leftarrow \mathbf{Q}^\transposed \mathbf{g}$
-		\State $l = \lambda_k + |a_k| - \code{kEps}, r = ||\mathbf{a}|| + \lambda_k + \code{kEps}$ \algorithmiccomment{Bracket root}
-		\State $\hat{\gamma} \leftarrow \code{bisect} (C, [l, r] \code{)}$ \algorithmiccomment{Find root of Eq.\ref{eq:characterisitc-poly} with bisection}
-		\State $\mathbf{f} \leftarrow \left(\frac{a_1}{\lambda_1 - \hat{\gamma}}, \frac{a_2}{\lambda_2 - \hat{\gamma}}, \frac{a_3}{\lambda_3 - \hat{\gamma}} \right)$
-		\State $\mathbf{n}^* \leftarrow \mathbf{Q}\mathbf{f}$ 
-		\State \Return $\mathbf{n}^*$
-		\EndFunction
-	\end{algorithmic}
-\end{algorithm}
-
-This algorithm is shown in Alg \ref{alg:solve-constrained-ev}, see \ref{sec:constrained-ev-problem-derivation} for a derivation with explanations.
-The eigen-decomposition can exploit the fact that the matrix is real and symmetric (but not positive semi-definite) for which faster algorithms are available (i.e.
-Eigen's \code{Eigen::SelfAdjointEigenSolver}).
-Note also that the vector $\mathbf{f}$ in line 8 is equal to $(\mathbf{\Lambda} - \hat{\gamma} \mathbf{I})^{-1} \mathbf{a}$.
-No actual matrix-inversion is required however since the matrix is diagonal and also with floating-point arithmetic it is numerically more stable/accurate to compute $a/b$ instead of $\frac{1}{b} a$.
-
-\subsection{Overall active-set solver}
-
-\begin{algorithm}[!ht]
-	\caption{Active-set method for minimizing the WLS-relaxation over rotations (\ref{eq:tls-rwls})}
-	\label{alg:min-wls-relax-rot}
-	\begin{algorithmic}[1]
-		\Require{Point cloud matrices $\mathbf{A}, \mathbf{B}$, node interval $(\mathbf{R_c}, n_r)$, computed WLS-relaxation weights vector $\mathbf{w}$}
-		\Ensure Solution to the optimization problem \ref{eq:tls-rwls}
-		\hypertarget{ref:min-wls-relax-rot2}{
-			\Function{\code{Minimize\_WLS\_Relax\_Rotation}}{$\mathbf{A}, \mathbf{B}, \mathbf{R_c}, n_r, \mathbf{w}$}}
-		\State $\mathbf{B} \leftarrow \sum_{i=1}^N w_i \mathbf{R}_c^\transposed \mathbf{a}_i\mathbf{b}_i^\transposed $ \algorithmiccomment{Compute cross-correlation matrix}
-		\State $\bar{\Delta \mathbf{R}} \leftarrow \code{svd\_rot(} \mathbf{B} \code{)}$  \algorithmiccomment{Solve unconstrained with SVD-algorithm}
-		\If{$d_{\angle}(\mathbf{I}, \Delta\mathbf{R}) \leq n_r$} \algorithmiccomment{Check if constraint is already satisfied}
-		\State \Return $\mathbf{R_c} \bar{\Delta \mathbf{R}}$
-		\EndIf
-		\State $\mathbf{A}, \mathbf{g} \leftarrow \code{get\_davenport\_matrices(} \mathbf{B}, n_r \code{)}$ \algorithmiccomment{Get  data-matrices (Eq. \ref{eq:davenport-data-matrices})}
-		\State $\mathbf{n}^* \leftarrow \code{solve\_constrained\_ev(} \mathbf{A}, \mathbf{g} \code{)}$ \algorithmiccomment{Solve for
-			rotation axis (Eq. \ref{eq:eq-constrained-prob-ev})}
-		\State $\bar{\Delta \mathbf{R}} \leftarrow \code{from\_axis\_angle(}\mathbf{n}^*, n_r \code{)}$ \algorithmiccomment{ Construct optimal rotation}
-		\State \Return $\mathbf{R_c} \bar{\Delta \mathbf{R}}$
-		\EndFunction
-	\end{algorithmic}
-\end{algorithm}
-
-Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is shown in Alg.
-\ref{alg:min-wls-relax-rot}.
-It is very efficient since the only computationally intensive operation is the computation of the $3 \times 3$ cross-correlation matrix $\mathbf{B}$ (line 13) that has linear runtime in the number of points.
+Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is is overall very efficient.
+The only computationally intensive operation is the computation of the $3 \times 3$ cross-correlation matrix $\mathbf{B}$ (line 13) that has linear runtime in the number of points.
 All the operations that follow are based on this $3 \times 3$ matrix and therefore have a constant runtime (i.e.
-independent of the number of points) and are in practice very fast.
-For example, the eigen-decomposition (line 3) takes only around $10 \mu s$ while the computation of the cross-correlation matrix requires aroundd $500 \mu s$ for 10k points.
-This is also because the cross-correlation matrix is re-used between the SVD-algorithm (line 3) and the algorithm for the equality-constraint case (line 7).
-The function \code{from\_axis\_angle} converts a given axis and angle to a rotation matrix using Rodrigues' formula.
-
+independent of the number of points).
+(For example, the eigen-decomposition (line 3) takes only around $10 \mu s$ while the computation of the cross-correlation matrix requires aroundd $500 \mu s$ for 10k points.)
 This solver is novel to our knowledge -- the problem of solving the Wahba-problem with inequality-constrained angle has not been proposed in the literature before.
-
-The solver for the equality-constraint case ($\code{solve\_constrained\_ev}$) is implemented from the derivation first done in \cite{10.1007/978-3-642-75536-1_57} and inspired by the \textit{Trust-Region Subproblem} (TRS) that arises in trust-region (local) optimization algorithms \cite{Adachi2017SolvingTT} such as Levenberg-Marquardt.
-For such TRS-problems, an active-set method has been proposed also in \cite{Rontsis2022} where it was generalized additionally for a minimum-norm constraints.

### 2025-04-21 sec/4_evaluation.tex
Shorten ineq solver

--- 2025-04-21 sec/4_evaluation.tex
@@
-\section{Experiments}
-In this section, we present experiments where we study the intersection graphs arising from the robust point cloud registration problem.
-We evaluate the runtime of state-of-the-art maximum clique finding algorithms.
+\section{Experimental evaluation}
+\label{sec:evaluation}
+In this section we evaluate our Branch and Bound solver that uses our novel convex relaxation on point cloud registration and related outlier robust problems. 
+We measure the required number of iterations as well as the per-iteration costs. We have implemented a simple version of a best-first Branch-and-Bound \cite{10.1007/978-3-642-37444-9_42}.
 
-\label{sec:evaluation}
+We also have validated the corecteness of the algorithm and it's implementation empirically: We did not find any instance in which it did not find the global minimum of the TLS problem.
+
 We perform the experiments on a desktop PC with an Intel i7 12700KF CPU and 32GiB of RAM.
-\subsection{Synthetic experiments}
 
@@
 
+\begin{figure}[!ht]
+	\centering
+	\begin{adjustbox}{width=1.\linewidth}
+		\includegraphics{figures/robust_wahba_runtime_rotation_error.pdf}
+	\end{adjustbox}
+	\caption{Der durchschnittliche Rotationsfehler in Abhängigkeit vom Anteil der Ausreißer: Unsere Methode bleibt robust bis zu 95\% Ausreißer, wie von der Truncated Least Squares Formulierung zu erwarten ist.}
+	\label{fig:rotation-error-over-outlier-rate}
+\end{figure}
+
+\begin{figure}[!ht]
+	\centering
+	\begin{adjustbox}{width=1.\linewidth}
+		\includegraphics{figures/robust_wahba_runtime_execution_time.pdf}
+	\end{adjustbox}
+	\caption{Die durchschnittliche Ausführungszeit von Branch-and-Bound in Abhängigkeit von der Gesamtzahl der Punkte $N$ des Optimierungsproblems.}
+	\label{fig:runtime}
+\end{figure}
+
 We generate synthetic point clouds by sampling inliers from a unit cube $s \cdot [-1, 1]^3$ with scale $s$.
@@
 As shown in Fig.
-\ref{fig:execution-time-over-outlier-rate}, the runtime is very low for $N =100$ and decreases until 90\% outlier-rate because the graph sparsity increases.
-Above 90\% outlier-rate, it increases because the number of visited cliques increases.
+\ref{fig:execution-time-over-outlier-rate}, the runtime is very low for $N =100$.
+
+The results of the rotation error as a function of the outlier rate are shown in Fig.
+\ref{fig:rotation-error-over-outlier-rate} is shown.
+The estimation remains correct with a small error of less than $1$ degree.
+We also investigated the scalability of our method over the total number of points $N$.
+As the results in Fig.
+\ref{fig:runtime} show, our approach can even be used for real-time applications with a runtime of less than 100 milliseconds.
 
@@
 \subsubsection{Validity of the global optimality certificate}
-Even though GNC is a heuristic, based on our case analysis (Sec. \ref{sec:branching-over-comb}) it seems likely that GNC finds global minimum under certain conditions.
-We therefore test on synthetic data whether the global optimality certificate is valid, i.e.
-whether at termination, the lower bound is less or equal than the TLS objective at the ground-truth solution.
+
+We test on synthetic data whether the global optimality certificate is valid, i.e. whether at termination, the lower bound is less or equal than the TLS objective at the ground-truth solution.
 We define the suboptimality $\eta$ of the solution based on the upper bound (UB), that is, the best known solution, and the lower bound (LB) \cite{9785843}
@@
 \end{figure}
-The results are shown in Fig.
-\ref{fig:cert-validity}, we clearly see that with a larger data scale, the likelihood of a valid certificate (and successful registration result) increases.
-While this observation alone could be explained solely by the solver being able to solve the problem more likely, we also see that the suboptimality increases for instances above 90\% outliers.
-This indicates that our solver is able to detect that some instances could not be solved to global optimality.
-Unfortunately, being only able to reject some and not all instances means our solver is currently neither globally optimal (being able to find the global minimum of every instance and outputting a valid lower bound on every instance), nor certifiable (outputting only a valid lower bound on every instance).
+
+TODO
 

### 2025-04-21 sec/3_tls_opt.tex
Polish t solver

@@
 
-To minimize the WLS-relaxation over the translation $\mathbf{t}$ for obtaining the lower bound over the ball with center $\mathbf{c}$ and radius $n_r$, we need to solve the following optimization problem:
+The problem of minimizing the WLS-relaxation over a ball $\{ \mathbf{t} : || \mathbf{c} - \mathbf{t} || \leq n_r \}$ of translations is:
+\begin{equation}
+	\label{eq:tls-twls-constrained}
+	\begin{aligned}
+		\argmin_{\mathbf{x} \in \Rone^3} \quad &\sum_{i=1}^{N} w_i ||\mathbf{a}_i - \mathbf{b}_i + \mathbf{c} - \mathbf{t}||_2^2 \\
+		\text{subject to} \quad  &||\mathbf{c} - \mathbf{t}||_2 \leq n_r
+	\end{aligned}
+\end{equation}
+
+The ball center $\mathbf{c}$ is the center of the BnB node that has a radius $n_r$. 
+The constraint $||\mathbf{c} - \mathbf{t}||_2 \leq n_r$ therefore ensures  that the found $\mathbf{t}$ is inside the BnB node.
+Note that compared to the definition of the convex relaxation \ref{eq:tls-wls-relaxation1}, two constant sums disappeared because they are constant (i.e.
+they do not depend on $\mathbf{t}$).
+We can write this problem equivalently as:
 \begin{equation}
@@
 \end{equation}
-
-where 
+with 
 \begin{equation}
 	\begin{aligned}
-		\mathbf{x} = \mathbf{c} - \mathbf{t}  \iff  \mathbf{t} = \mathbf{c} - \mathbf{x}\\
-		\mathbf{d}_i = \mathbf{a}_i - \mathbf{b}_i + \mathbf{c}
+		\mathbf{x} = \mathbf{c} - \mathbf{t}, \quad \mathbf{d}_i = \mathbf{a}_i - \mathbf{b}_i + \mathbf{c}
 	\end{aligned}
@@
 
-The constraint $||\mathbf{x}||_2 \leq n_r$ ensures that the found $\mathbf{x}$ is inside the BnB node.
-The other sums of the convex relaxation \ref{eq:tls-wls-relaxation1} disappeared because they are constant (i.e.
-they do not depend on $\mathbf{x}$).
-This problem is a convex QCQP and therefore solvable by any standard QCQP solver.
-However, we show that the solution can be obtained even more easily in closed form, which in practice results in a much faster algorithm.
-
-We first form the Lagrangian to account for the constraint (which is squared):
-
+This problem can be solved easily in closed form.
+We first form the Lagrangian to account for the constraint:
 \begin{equation}

### 2025-04-21 sec/3_tls_opt.tex
Polish rot solve

@@
 
-The problem of minimizing the WLS-relaxation over a ball $\{ \mathbf{t} : || \mathbf{c} - \mathbf{t} || \leq n_r \}$ of translations is:
+The problem of minimizing the WLS-relaxation over a ball $\{ \mathbf{t} \in \Rthree : || \mathbf{c} - \mathbf{t} || \leq n_r \}$ of translations is:
 \begin{equation}
@@
 	\begin{aligned}
-		\argmin_{\mathbf{x} \in \Rone^3} \quad &\sum_{i=1}^{N} w_i ||\mathbf{a}_i - \mathbf{b}_i + \mathbf{c} - \mathbf{t}||_2^2 \\
+		\argmin_{\mathbf{t} \in \Rone^3} \quad &\sum_{i=1}^{N} w_i ||\mathbf{a}_i - \mathbf{b}_i + \mathbf{c} - \mathbf{t}||_2^2 \\
 		\text{subject to} \quad  &||\mathbf{c} - \mathbf{t}||_2 \leq n_r
@@
 
-This optimization problem can be understood as a special case of the Trust Region Problem for which strong duality holds \cite[Ch.
-5.2, p.229]{Boyd_Vandenberghe_2004} as well as a regularized linear regression problem.
+This optimization problem is a special case of the Trust Region Problem for which strong duality holds \cite[Ch. 5.2, p.229]{Boyd_Vandenberghe_2004}.
 
@@
 
-To minimize the convex relaxation over rotations and obtain a lower bound for given BnB-node that is a ball with center $\mathbf{R_c}$ and (geodesic) radius $n_r$, we need to solve the following optimization problem:
+To problem of minimizing the WLS relaxation over a geodesic ball of rotations $\{\mathbf{R} \in \SOthree:  d_{\angle}(\mathbf{R}_c, \Delta\mathbf{R}) \leq n_r\}$ is:
 
+\begin{equation}
+	\label{eq:tls-rwls}
+	\begin{aligned}
+		\argmin_{\mathbf{R} \in \SOthree}  \quad &\sum_{i=1}^{N} w_i ||\mathbf{R}\mathbf{p}_i - \mathbf{q}_i ||_2^2 \\
+		\text{subject to} \quad &d_{\angle}(\mathbf{R}_c, \Delta\mathbf{R}) \leq n_r
+	\end{aligned}
+\end{equation}
+
+where $\mathbf{R_c}$ is the ball center and $n_r$ is the radius, as in the translation case. $d_{\angle}(\cdot)$ is the geodesic (also called angular) distance.
+We can write this problem equivalently (due to the bi-invariance of the metric over $\SOthree$) as:
 \begin{equation}
@@
 
-Where $\Delta\mathbf{R}$ is the rotation relative to the node's center:
+Where $\Delta\mathbf{R}$ is the rotation relative to the node's center and $\mathbf{I}$ is the identity rotation:
 
@@
 	\begin{aligned}
-		\Delta\mathbf{R} = \mathbf{R_c}^\transposed\mathbf{R} \iff \mathbf{R_c}\Delta\mathbf{R} = \mathbf{R}
+		\Delta\mathbf{R} = \mathbf{R_c}^\transposed\mathbf{R} 
 	\end{aligned}
@@
 
-The unconstrained version of this problem is known as the \textit{Wahba-problem} for which a simple SVD-algorithm exists, known as the Kabsch-Umeyama algorithm \cite{Kabsch-1978-Point-set-alignment} \cite{Least-squares-estimation-point-sets-Umeyama-1991} \cite{Lawrence2019APA}.
-But this problem is modified and has in addition an inequality-constraint.
-In the following, we develop a custom solver that solves this problem very efficiently.
+The unconstrained version of this problem is known as the \textit{Wahba-problem} for which a simple SVD-algorithm exists \cite{Kabsch-1978-Point-set-alignment} \cite{Lawrence2019APA} \cite{Least-squares-estimation-point-sets-Umeyama-1991}.
+This problem however is modified and has additionally an inequality-constraint.
+We show that this problem can be solved computationally very efficient using a custom active-set solver.
 
-The first idea is to just try to solve the unconstrained version and check whether it satisfies the inequality constraint (i.e. the solution is inside the ball). 
-We can do that, but unfortunately the solution does not always satisfy the inequality constraint.
-We gain however an important insight: If the solution satisfies the constraint and lies therefore in the spherical region, we have already found the optimum of the constrained problem.
-If however the optimum of the unconstrained problem is outside of the region, the optimal solution to the constrained problem has to lie on the border of the region.
-It then has to satisfy the constraint as an equality, i.e.
-$ d_{\angle}(\mathbf{I}, \Delta\mathbf{R})  = n_r$ \cite[Ch.15 p.424-427]{Numerical-Optimization-Nocedal-Wright}.
-This two-stage method for solving inequality-constrained nonlinear problems is known as the \textit{active-set} method \cite[Ch.
-16.5, p.467]{Numerical-Optimization-Nocedal-Wright}, an inequality constraint is set to be active if the solution satisfies it as an equality.
+The idea of the active set method is based on the observation that every inequality constraint can either be ignored or is satisfied as an equality (also called complementary slackness condition) \cite[Ch.15 p.424-427]{Numerical-Optimization-Nocedal-Wright}.
+We therefore first solve the problem using the SVD algorithm and thereby ignoring the constraint.
+Then, we simply check whether the constraint is satisfied.
+If yes, we already have found the solution.
+If not, we know that we need to find a 3D rotation that has a rotation angle equal to $n_r$.
+By the Euler theorem, every 3D rotation can be described as a rotation around a single axis and an single angle.
+Because we know that the rotation angle needs to be $n_r$, we therefore only need to find the rotation axis.
 
-Using this active-set approach, we can effectively solve this inequality-constrained optimization problem, since we already have the SVD-algorithm for the unconstrained case.
-And for equality-constrained case, we make the observation that by Euler's theorem that says that every 3D-rotation can be separated into a n axis and angle, we only have to optimize over the axis when the angle is equality-constrained and therefore fixed.
-
-As an algorithm outline, we therefore first solve the unconstrained problem with the SVD-algorithm and check whether it satisfies the inequality-constraint.
-If not, we solve for the equality-constrained problem where we seek for the rotation that has the rotation angle $n_r$, that is more difficult case.
-
-\subsection{Solving the equality-constrained case}
-
-If the inequality-constraint is not satisfied after the first unconstrained solve, we need to solve the problem with equality-constraint, that is the following optimization problem:
+\subsubsection{Solving the equality-constrained case}
+In the equality-constrained case of the active-set solver, we need to find a 3D rotation with a rotation angle of exactly $n_r$:
 
@@
 
-We show that this optimization problem is equivalent to the following problem that finds the optimal axis of rotation $\mathbf{n}^*$:
+We show that this optimization problem is equivalent to the following problem over the axis of rotation:
 
@@
 
-with the data matrices $\mathbf{A}$ and $\mathbf{g}$ that depend on the points $\mathbf{a}_i$ and $\mathbf{b}_i$ and the angle constraint $n_r$.
+The data matrices $\mathbf{A}$ and $\mathbf{g}$ depend on the points $\mathbf{a}_i$ and $\mathbf{b}_i$ and the angle constraint $n_r$.
 A definition is given in the Appendix \ref{thm:constrained-wahba-davenport}.
@@
 To solve this non-convex optimization problem \ref{eq:eq-constrained-prob-ev} over the sphere, we use the algorithm proposed in \cite{10.1007/978-3-642-75536-1_57}.
-We perform overall two steps: We first need to compute the eigenvalue decomposition  $\mathbf{A} = \mathbf{Q}\mathbf{\Lambda}\mathbf{Q}^\transposed$ and then need to find the rightmost root of a certain rational function (characteristic polynomial).
+We perform overall two steps: We first need to compute the eigenvalue decomposition  $\mathbf{A} = \mathbf{Q}\mathbf{\Lambda}\mathbf{Q}^\transposed$ and then need to find the rightmost root of a certain rational function (a characteristic polynomial).
 This can be done in practice very fast, yielding a constant time algorithm.
@@
 Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is is overall very efficient.
-The only computationally intensive operation is the computation of the $3 \times 3$ cross-correlation matrix $\mathbf{B}$ (line 13) that has linear runtime in the number of points.
+The only computationally intensive operation is computing the $3 \times 3$ cross-correlation matrix that is needed first by the SVD algorithm.
+It has a linear runtime in the number of points.
 All the operations that follow are based on this $3 \times 3$ matrix and therefore have a constant runtime (i.e.

### 2025-04-21 sec/X_suppl.tex
Shorten ineq solver
Cleaned up equality constrained case proof

@@
 \begin{theorem}
-
+	\label{thm:constrained-wahba-davenport}
 	The least-squares rotation estimation problem 
@@
 	\end{equation}
-	where the angle of rotation is constrained to $n_r$ is equivalent to the following optimization problem over the rotation axis $\mathbf{n}$:
+	where the angle of rotation is constrained equal to $n_r$ is equivalent to the following optimization problem over the rotation axis $\mathbf{n}$:
 	\begin{equation}
@@
 		\begin{aligned}
+			\mathbf{A} &= \sin^2 \left(\frac{n_r}{2} \right) \mathbf{C}, \quad \mathbf{g} =  \sin \left(\frac{n_r}{2} \right) \cos \left(\frac{n_r}{2} \right) \mathbf{z}\\
+			\mathbf{C} &= \mathbf{B} + \mathbf{B}^\transposed\\
 			\mathbf{B} &= \sum_{i=1}^N w_i \mathbf{a}_i\mathbf{b}_i^\transposed \in \Rone^{3 \times 3}, \quad
-			\mathbf{z} = \sum_{i=1}^N w_i [\mathbf{a}_i]_{\times}\mathbf{b}_i \in \Rone^{3}, \quad \mathbf{C} = \mathbf{B} + \mathbf{B}^\transposed\\
-			\mathbf{A} &= \sin^2 \left(\frac{n_r}{2} \right) \mathbf{C}, \quad \mathbf{g} =  \sin \left(\frac{n_r}{2} \right) \cos \left(\frac{n_r}{2} \right) \mathbf{z}
+			\mathbf{z} = \sum_{i=1}^N w_i [\mathbf{a}_i]_{\times}\mathbf{b}_i \in \Rone^{3}
 		\end{aligned}
@@
 
-	We first separate the optimization over the rotation axis and angle.
-For this, we use the quaternion representation of \ref{eq:tls-rwls-a}.
-One can show that the problem \ref{eq:tls-rwls-a} is equivalent to:
-
+	The problem \ref{eq:tls-rwls-a} without the equality constraint (i.e. the Wahba problem) can be stated equivalently using quaternions as:
 	\begin{equation}
@@
 	\end{equation}
-
-	where $\mathbf{q}$ is a 4-dimensional vector representing the quaternion of rotation and $\mathbf{K} \in \Rone^{4\times4}$ is a data matrix.
-This optimization problem is known as the \textit{Davenport's q-method}.
-The quaternion can be written in axis-angle form as:
-
-	\begin{equation}
-		\label{eq:q-axis-angle}
-		\begin{aligned}
-			\mathbf{q} = \left( \sin \left( \frac{\theta}{2} \right) \mathbf{n}, \cos \left(\frac{\theta}{2} \right) \right)^\transposed
-		\end{aligned}
-	\end{equation}
-	i.e.
-the first three elements are the unit-vector $\mathbf{n} \in \Rthree, ||\mathbf{n}|| = 1$ of the axis of rotation multiplied by $\sin \left( \frac{\theta}{2} \right)$, while the fourth element is just $\cos \left(\frac{\theta}{2} \right)$ and $\theta$ is the angle of rotation.
+	where $\mathbf{q}$ is the quaternion and $\mathbf{K} \in \Rone^{4\times4}$ is a data matrix.
 	The data matrix $\mathbf{K}$ has the following block-structure:
@@
 
-	We can rewrite the objective function $\mathbf{q}^\transposed \mathbf{K} \mathbf{q}$  of \ref{eq:davenports-q-problem} using the blocks of the block-matrix $\mathbf{K}$:
+	where the matrices $\mathbf{C}$ and $\mathbf{z}$ are defined as in the theorem \ref{eq:ev-constr-data-mat-a}. 
+	We can separate now the optimization over the rotation axis and angle using the following definition:
+
+	\begin{equation}
+		\label{eq:q-axis-angle}
+		\begin{aligned}
+			\mathbf{q} = \left( \sin \left( \frac{\theta}{2} \right) \mathbf{n}, \cos \left(\frac{\theta}{2} \right) \right)^\transposed
+		\end{aligned}
+	\end{equation}
+	where $\mathbf{n}$ is a unit-vector that is the axis of rotation and $\theta$ is the angle of rotation.
+	We rewrite the objective function $\mathbf{q}^\transposed \mathbf{K} \mathbf{q}$  of \ref{eq:davenports-q-problem} using the blocks of the block-matrix $\mathbf{K}$:
 
@@
 
-	We now have a scalar term where the axis $\mathbf{n}$ and the angle $\theta$ occur separately and with this we can formulate the optimization problem over the axis.
-By setting $\theta = n_r$ and solving only over $\mathbf{n}$ we can therefore solve the equality-constrained problem \ref{eq:tls-rwls-a}.
-
-	Since the term $ \cos^2 \left(\frac{\theta}{2} \right) \trace(\mathbf{C})$ is constant in Eq. \ref{eq:dv-q-rew}, it does no affect the minimizer and we will leave it out. 
-	By defining the data-matrices as in \ref{eq:ev-constr-data-mat-a}, we immediatelly see that the problem 
-	\ref{eq:rot-est-constrained-qcqp} is equivalent to Davenports Q-problem \ref{eq:davenports-q-problem} and therefore equivalent to \ref{eq:tls-rwls-a}, concluding the proof.
+	Now the axis $\mathbf{n}$ and the angle $\theta$ occur separately. 
+	By the equality constraint, $\theta$ is already given as $\theta = n_r$.
+Then, the term $ \cos^2 \left(\frac{\theta}{2} \right) \trace(\mathbf{C})$ in Eq.
+\ref{eq:dv-q-rew} is constant and does not affect the minimizer, we will therefore omit it.
+	 By defining the data-matrices $\mathbf{A}$ and $\mathbf{g}$ as in the theorem \ref{eq:ev-constr-data-mat-a}, we see that the objective function of Eq.
+\ref{eq:rot-est-constrained-qcqp} is equivalent to Davenports' problem \ref{eq:davenports-q-problem} and therefore equivalent to \ref{eq:tls-rwls-a}, concluding the proof.
 \end{proof}

### 2025-04-21 sec/0_abstract.tex
Some polishing

@@
 with global optimality guarantee, remains challenging.
-Approaches using semidefinite relaxations are not well scalable albeit running in polynomial time and instead have a runtime on the order of hundreds of seconds for 100 points.
-O
+Approaches using semidefinite convex relaxations albeit running in polynomial time do not scale well and instead have a runtime on the order of hundreds of seconds for 100 points.
 
 We propose a new convex relaxation that runs in linear time and can be used in combination with Branch-and-Bound.
-We present results on synchetic data for the rotation-only (Wahba) problem and translation problem separately.
+It is two orders of magnitude faster when solving the rotation-only problem compared to state-of-the-art SDP solver STRIDE.
+We present results on synthetic data for the rotation-only (Wahba) problem and translation problem separately.
 

### 2025-04-21 sec/3_tls_opt.tex
Some polishing

@@
 The idea of the WLS-relaxation is the observation that this function is only non-convex because of the truncation -- the residuals are convex (parabolas).
-Also, sums of convex functions are convex (this is the reason why ordinary least squares is a convex problem), so we can focus on making each term convex separately.
+Since sums of convex functions are convex, we can focus on relaxing each term separately.
 
@@
 
-However, some additional special case handling is required to ensure that this relaxation becomes tight when computing over (small) intervals where the TLS objective is already convex.
+Some additional case handling is required however to ensure that this relaxation becomes tight.
 This is crucial to ensure that BnB converges in a finite number of iterations.
-First, the reweighting should only be applied to residuals where $r_{min} < \epsilon^2$, i.e., residuals that are not already outliers over the entire interval.
-Similarly, residuals where $r_{max} < \epsilon^2$ are inliers over the entire interval and are therefore already convex and do not need to be reweighted.
+First, all residuals for which  $r_{min} \geq \epsilon^2$ holds are outliers over the entire interval and therefore constant.
+Similarly, residuals where $r_{max} \leq \epsilon^2$ holds are inliers over the entire interval and do not need to be reweighted.
 
-After considering these special cases, we can show that $\text{WLS}_{r}$ is a convex relaxation of \ref{eq:tls-scalar-for-wls} over a given interval:
+After considering these special cases, we can show that $\text{WLS}_{relax}$ is a convex relaxation of \ref{eq:tls-scalar-for-wls} over a given interval:
 \begin{equation}
@@
 	\begin{aligned}	
-		\text{WLS}_{r} &= 
+		\text{WLS}_{relax} &= 
 		\sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right) + \sum_{i=1}^{N} o_i\epsilon^2\\
@@
 The unconstrained version of this problem is known as the \textit{Wahba-problem} for which a simple SVD-algorithm exists \cite{Kabsch-1978-Point-set-alignment} \cite{Lawrence2019APA} \cite{Least-squares-estimation-point-sets-Umeyama-1991}.
-This problem however is modified and has additionally an inequality-constraint.
-We show that this problem can be solved computationally very efficient using a custom active-set solver.
+Our problem however has additionally an inequality-constraint.
+It still can be solved computationally very efficient using a custom active-set solver.
 
 The idea of the active set method is based on the observation that every inequality constraint can either be ignored or is satisfied as an equality (also called complementary slackness condition) \cite[Ch.15 p.424-427]{Numerical-Optimization-Nocedal-Wright}.
-We therefore first solve the problem using the SVD algorithm and thereby ignoring the constraint.
+We therefore first solve the problem using the SVD algorithm and thereby ignore the constraint.
 Then, we simply check whether the constraint is satisfied.
@@
 Because we know that the rotation angle needs to be $n_r$, we therefore only need to find the rotation axis.
+
+Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is is overall very efficient.
+The only computationally intensive operation is computing the $3 \times 3$ cross-correlation matrix that is needed first by the SVD algorithm.
+It has a linear runtime in the number of points.
+All the operations that follow are based on this $3 \times 3$ matrix and therefore have a constant runtime (i.e.
+independent of the number of points).
+This solver is novel to our knowledge -- the problem of solving the Wahba-problem with inequality-constrained angle has not been proposed in the literature before.
 
@@
 This can be done in practice very fast, yielding a constant time algorithm.
-
-Our active-set solver for finding the global minimum of the WLS-relaxation over rotations is is overall very efficient.
-The only computationally intensive operation is computing the $3 \times 3$ cross-correlation matrix that is needed first by the SVD algorithm.
-It has a linear runtime in the number of points.
-All the operations that follow are based on this $3 \times 3$ matrix and therefore have a constant runtime (i.e.
-independent of the number of points).
-(For example, the eigen-decomposition (line 3) takes only around $10 \mu s$ while the computation of the cross-correlation matrix requires aroundd $500 \mu s$ for 10k points.)
-This solver is novel to our knowledge -- the problem of solving the Wahba-problem with inequality-constrained angle has not been proposed in the literature before.

### 2025-04-21 sec/4_evaluation.tex
Some polishing

@@
 In this section we evaluate our Branch and Bound solver that uses our novel convex relaxation on point cloud registration and related outlier robust problems. 
-We measure the required number of iterations as well as the per-iteration costs. We have implemented a simple version of a best-first Branch-and-Bound \cite{10.1007/978-3-642-37444-9_42}.
+We measure the required number of iterations as well as the per-iteration costs.
+For this we have implemented a simple version of a best-first Branch-and-Bound as in \cite{10.1007/978-3-642-37444-9_42}.
 
@@
 	\end{adjustbox}
-	\caption{Der durchschnittliche Rotationsfehler in Abhängigkeit vom Anteil der Ausreißer: Unsere Methode bleibt robust bis zu 95\% Ausreißer, wie von der Truncated Least Squares Formulierung zu erwarten ist.}
+	\caption{Rotation-only experiments: The average rotation error over the ourlier rate: The Truncated Least Squares method is robust up to 95\% outliers.}
 	\label{fig:rotation-error-over-outlier-rate}
@@
 	\end{adjustbox}
-	\caption{Die durchschnittliche Ausführungszeit von Branch-and-Bound in Abhängigkeit von der Gesamtzahl der Punkte $N$ des Optimierungsproblems.}
+	\caption{Average runtime of our Branch-and-Bound solver depending on the total amount of point correspondences $N$.}
 	\label{fig:runtime}

### 2025-05-05 sec/3_tls_opt.tex
Writing

@@
 
+\subsubsection{Translation}
 The minimum $r_{min}^i$ and maximum $r_{max}^i$ of the residual $r_i =||\mathbf{a}_i - \mathbf{b}_i + \mathbf{t}||_2^2$ (translation) over a ball with center $\mathbf{c}$ and radius $n_r$ are: (see \ref{proof:tls-relax-minmax-r} for a proof):
@@
 \end{equation}
+
+\subsubsection{Rotation}
 
@@
 
+\subsubsection{Rotation and Translation (3D-transform)}
+
+To solve the full 6DoF problem, we will need the residual interval over a translation ball and \textit{any} rotation, later it  will become aparent why no rotation ball is needed.
+
+The minimum over all translations $\vt \in B^t(\vc) := \{\vt \in \Rthree: \norm{\vc - \vt}_2 \leq n_r\}$ for all $\vp, \vq \in \Rthree$ is
+\begin{equation}
+	\begin{aligned}
+		&\min_{\MR \in \SOthree, \vt \in B^t(\vc)} \normsq{\MR \vp - \vq + \vt}\\
+		&= \max \left(  0, | \norm{\vp} - \norm{\vq + \vc} | - n_r \right)^2
+	\end{aligned}
+\end{equation}
+
 \section{Minimizing the WLS convex relaxation}
@@
 The only thing we need to consider is that we seek the minimum that is inside the Branch and Bound node.
-We therefore need to introduce an inequality constraint into the problem.
-This constrained problem can be solved can still be solved very efficiently by utilizing custom active-set solvers that are only slightly more expensive than solving the least-squares problem.
+We therefore need to introduce an inequality constraint into the minimization problem.
+This constrained problem still can be solved very efficiently by utilizing custom active-set solvers, they are only slightly more expensive than solving the least-squares problem.
 

### 2025-05-05 sec/4_evaluation.tex
Conc

@@
 \label{sec:evaluation}
-In this section we evaluate our Branch and Bound solver that uses our novel convex relaxation on point cloud registration and related outlier robust problems. 
-We measure the required number of iterations as well as the per-iteration costs.
-For this we have implemented a simple version of a best-first Branch-and-Bound as in \cite{10.1007/978-3-642-37444-9_42}.
+In this section we evaluate a Branch and Bound solver that uses our novel convex relaxation on point cloud registration and related outlier robust problems.
+We have implemented a basic version of BnB using best-first search.
+To branch over the spherical space of 3D rotations, we use rotation vector approach as in \cite{10.1007/978-3-642-37444-9_42}.
 
-We also have validated the corecteness of the algorithm and it's implementation empirically: We did not find any instance in which it did not find the global minimum of the TLS problem.
+We also have validated the correctness of the algorithm and it's implementation empirically: We did not find any instance in which the lower bound was invalid, i.e.
+higher than the ground-truth global minimum.
 
 We perform the experiments on a desktop PC with an Intel i7 12700KF CPU and 32GiB of RAM.
+
+\subsection{Translation experiments}
+We evaluate solving the translation and the rotation problem separately.
 
@@
 	\centering
-	\includegraphics[width=0.49\linewidth]{figures/execution_time_over_outlier_rate_boxplot}
-	\includegraphics[width=0.49\linewidth]{figures/rotation_error_over_outlier_rate.pdf}
+%	\includegraphics[width=0.49\linewidth]{figures/execution_time_over_outlier_rate_boxplot}
+%	\includegraphics[width=0.49\linewidth]{figures/rotation_error_over_outlier_rate.pdf}
 	\caption{Execution time and rotation error over different outlier rates with $N=100$ on synthetic data.}
@@
 	\centering
-	\begin{adjustbox}{width=1.\linewidth}
-		\includegraphics{figures/robust_wahba_runtime_rotation_error.pdf}
-	\end{adjustbox}
-	\caption{Rotation-only experiments: The average rotation error over the ourlier rate: The Truncated Least Squares method is robust up to 95\% outliers.}
+	\caption{Rotation-only experiments: The average rotation error over the outlier rate: The Truncated Least Squares method is robust up to 95\% outliers.}
 	\label{fig:rotation-error-over-outlier-rate}
@@
 	\centering
-	\begin{adjustbox}{width=1.\linewidth}
-		\includegraphics{figures/robust_wahba_runtime_execution_time.pdf}
-	\end{adjustbox}
 	\caption{Average runtime of our Branch-and-Bound solver depending on the total amount of point correspondences $N$.}
@@
 	\centering
-	\includegraphics[width=0.9\linewidth]{figures/scalability-synth}
+	%\includegraphics[width=0.9\linewidth]{figures/scalability-synth}
 	\caption{Scalability on synthetic data, outlier rate $75\%$}
@@
 
+\subsection{Rotation experiments}
+We solve the rotation-only truncated least squares problem with the same approach as the translation problem: We use the same BnB solver and convex relaxation, only the computation of the residual intervals changes, as well as the minimization of the relaxation function.
+
 \subsubsection{Validity of the global optimality certificate}
-
 We test on synthetic data whether the global optimality certificate is valid, i.e. whether at termination, the lower bound is less or equal than the TLS objective at the ground-truth solution.
@@
 \end{equation}
-We test with $N=100$, $\epsilon=0.5$ and vary the data scale $s$ from 3 to 100. We test for each scale outlier rates of $0\%, 30\%, 50\%, 90\%, 96\%$, and do 1000 trials for each combination.
+We test with $N=100$, $\epsilon=0.5$ and vary the data scale $s$ from 3 to 100.
+We test for each scale variation the outlier rates $0\%, 30\%, 50\%, 90\%, 96\%$.
+We do 1000 trials for each such parameter set.
+
 \begin{figure}[!ht]
 	\centering
-	\includegraphics[width=0.49\linewidth]{figures/certificate_validity_vs_scale}
-	\includegraphics[width=0.49\linewidth]{figures/eta_subopt_over_outlier_rate}
+	%\includegraphics[width=0.49\linewidth]{figures/certificate_validity_vs_scale}
+	%\includegraphics[width=0.49\linewidth]{figures/eta_subopt_over_outlier_rate}
 	\caption{Amount of instances with valid global optimality certificate (i.e. valid suboptimality) on synthetic data, including achieved $\eta$-suboptimality.}
@@
 	\centering
-	\includegraphics[width=0.49\linewidth]{figures/subopt_stride}
-	\includegraphics[width=0.49\linewidth]{figures/runtime_stride}
+	%\includegraphics[width=0.49\linewidth]{figures/subopt_stride}
+	%\includegraphics[width=0.49\linewidth]{figures/runtime_stride}
 	\caption{The reached suboptimality and the runtime using STRIDE over the outlier rate on synthetic data, $N=30$.}
@@
 	\centering
-	\includegraphics[width=0.9\linewidth]{figures/rotation_error_stride}
+	%\includegraphics[width=0.9\linewidth]{figures/rotation_error_stride}
 	\caption{Rotation error using STRIDE.}
@@
 The runtime of STRIDE is even with as few as 30 points in the order of $10^3$ seconds.
-STRIDE achieves a similar rotation error compared to our solver in instances that could be solved (Fig.
+In instances that STRIDE could solve (i.e.
+where the relaxation is tight), it achieves a similar rotation error compared to our solver (Fig.
 \ref{fig:stride-rot-err}).
 Note that the rotation error is meaningless for instances that STRIDE could not solve (i.e.
-above 80\%), since it is always able to detect them.
+above 80\% outliers).
 
@@
 \input{sec/eval-3DMatch-table}
-We evaluate our solver on the 3DMatch dataset \cite{zeng20163dmatch} 
-following \cite{SC2-PCR-Chen-2022-CVPR, zhang20233d} on all scenes.
-We evaluate the residual rotation (RE) and translation (TE) and the registration recall (RR).
-A registration is defined as successful if RE $\leq 15$ degrees and TE $\leq 30$ cm.
-We sub-sample the correspondeces to 500 and set the inlier-threshold to $\epsilon = 0.6$.
+We evaluated our solver on the 3DMatch dataset \cite{zeng20163dmatch} 
+following the same protocol as in \cite{SC2-PCR-Chen-2022-CVPR, zhang20233d}.
+We use all scenes and evaluate the residual rotation error (RE) and translation error (TE) as well as the registration recall (RR).
+A registration is defined as successful if RE $\leq 15$ degrees and TE $\leq 30$ cm holds.
+We sub-sample the correspondences to 500 and set the inlier-threshold to $\epsilon = 0.6$.
 For the other methods, we report the results from \cite{zhang20233d} since the test setup is identical.

### 2025-05-05 sec/5_conclusion.tex
Conc

@@
 
-In this paper, we presented a general approach of minimizing the Truncated Least Squares (TLS) objective motivated by the geometry of the problem.
-We showed how intersection graphs allow for an elegant formulation of the combinatorial problem and how the global minimum can be found among the maximal cliques.
-We showed that surprisingly, the previously studied compatibility graphs motivated by problem-specific geometric observations lead to these special classes of graphs, namely the intersection graphs.
-We give a sufficient condition under which the global optimum is a maximal clique and discuss cases where subcliques may be the global minimum.
-Unfortunately, we were not able to fully prove that we can detect all conditions under which subcliques may be the global optimum.
-We think that a better understanding of the $\SEthree$-manifolds and their geometry including their combinatorial properties could lead to efficient algorithms for detecting such cases.
+In this paper, we presented a general approach of minimizing the Truncated Least Squares (TLS) objective function 
+using a fast linear time convex relaxation and Branch-and Bound.
+Our convex relaxation is general and can be applied for every model such as translation, rotation and rigid transform as long as residual intervals can be computed.
+We gave a full proof
+for the correctness of the relaxation, which is the main part of the proof of global optimality of a BnB algorithm.
+Our evaluation shows that the cost to compute and minimize the relaxation is in the single milliseconds range, allowing to solve the 
+translation and rotation-only problem two orders of magnitude faster than with the SoTA SDP-based method STRIDE.
+More work is however needed to solve the full 6DoF rigid transformation problem.
+Altought our relaxation can be applied here as well, branching over the 6D space is overall too slow.
+Further work in this direction will focus on contracting the rotation search space given a translation space, so that the search is not excessively inefficient.

### 2025-05-05 sec/3_tls_opt.tex
Conc
Polishc

@@
 	\centering
-	\includegraphics[width=0.6\linewidth]{figures/wls_relax_idea_with_relax}
-	\caption{The idea of the WLS convex relaxation: We compute the maximum $r_{max}$ of each convex term (orange) over the given interval (blue) and then reweight it such that it remains below $\epsilon^2$ over this interval and is thus not truncated over the entire interval (green).}
+	\caption{The idea of the WLS convex relaxation: We compute the maximum $r_{max}$ of each convex term (orange) over the given interval (blue) and then re-weight it such that it remains below $\epsilon^2$ over this interval and is thus not truncated over the entire interval (green).}
 	\label{fig:wlsrelaxideawithrelax}
@@
 
-Now the Karush-Kuhn-Tucker (KKT) conditions require $\lambda \geq 0$ and $\lambda (
+Now the Karush-Kuhn-Tucker (KKT) condition of complementary slackness requires $\lambda \geq 0$ and $\lambda (
 \mathbf{x}^\transposed \mathbf{x} - n_r^2) = 0$ in addition to satisfying the constraint, i.e.
 $|| \mathbf{x}||_2 \leq n_r$.
-It follows that in practice we obtain the optimum by first computing $\mathbf{x}$ with Eq.
+It follows that we can obtain the optimum by first computing $\mathbf{x}$ with Eq.
 \ref{eq:trans-qcqp-lagrangian}, assuming $\lambda = 0$.
 Then we check whether the inequality constraint is satisfied, i.e.
-whether $|| \mathbf{x}||_2 \leq n_r$ holds, and if not, we simply normalise $\mathbf{x}$ so that $|| \mathbf{x}||_2 = n_r$, yielding the optimal solution.
-Intuitively, from Eq.
-\ref{eq:trans-qcqp-lagrangian} we already see that the direction of the vector does not change regardless of the optimal Lagrange multiplier $\lambda$ since it is only a scalar factor.
+whether $|| \mathbf{x}||_2 \leq n_r$ holds, and if not, we simply normalize $\mathbf{x}$ so that it has the norm $n_r$.
+In other words, we see already from Eq.
+\ref{eq:trans-qcqp-lagrangian} that the Lagrange multiplier does not change the direction of $\mathbf{x}$, only its length.
 
@@
 To solve this non-convex optimization problem \ref{eq:eq-constrained-prob-ev} over the sphere, we use the algorithm proposed in \cite{10.1007/978-3-642-75536-1_57}.
-We perform overall two steps: We first need to compute the eigenvalue decomposition  $\mathbf{A} = \mathbf{Q}\mathbf{\Lambda}\mathbf{Q}^\transposed$ and then need to find the rightmost root of a certain rational function (a characteristic polynomial).
+We perform overall two steps: We first compute the eigenvalue decomposition $\mathbf{A} = \mathbf{Q}\mathbf{\Lambda}\mathbf{Q}^\transposed$ and then find the rightmost root of a certain rational function (a characteristic polynomial).
 This can be done in practice very fast, yielding a constant time algorithm.

### 2025-05-10 sec/3_tls_opt.tex
Moving TLS relax

@@
 
-\section{Wheighted Least Squares convex relaxation}
+\section{WLS Relaxation: A simple linear time relaxation}
 
-In this section we propose a novel convex relaxation for the Truncated Least Squares problem that can be computed in linear time.
+In this section we propose a novel convex relaxation for the Truncated Least Squares problem that can be computed in linear time and can be minimized by solving a simple weighted least squares problem.
 
@@
 In the following, we will refer to the i-th residual term $r(\mathbf{x}, \mathbf{x_i})^2$ simply as $r_i$.
-The idea of the WLS-relaxation is the observation that this function is only non-convex because of the truncation -- the residuals are convex (parabolas).
+The idea of the WLS-relaxation is the observation that this function is only non-convex because of the truncation -- the residuals itself are convex.
 Since sums of convex functions are convex, we can focus on relaxing each term separately.
 
-For this, we first find the minimum and maximum values of each residual $r_i$ is over the given interval, denoted as $r_{min}^i$ and $r_{max}^i$ respectively.
-The key geometric idea of our relaxation is to reweight these residuals such that after reweighting, their maximum becomes $\epsilon^2$ and thus they are not affected by truncation (Fig.
+For this, we first find the minimum and maximum values of each residual $r_i$ is over a given interval, denoted as $r_{min}^i$ and $r_{max}^i$ respectively.
+We then re-weight these residuals such that their maximum becomes $\epsilon^2$ and thus they are not truncated over the entire interval (Fig.
 \ref{fig:wlsrelaxideawithrelax}).
@@
 
-In the following we show how $r_{min}$ and $r_{max}$ can easily be computed analytically for different models (i.e. translation and rotation), given spherical intervals, i.e. balls.
+In the following we show how $r_{min}$ and $r_{max}$ can easily be computed analytically for different models (i.e. translation and rotation), given spherical intervals (balls).
 
@@
 
-This problem can be solved easily in closed form.
-We first form the Lagrangian to account for the constraint:
+This problem can be solved easily in closed form, obtaining the optimal solution $\mathbf{x}^*$ with:
+
 \begin{equation}
+	\label{eq:trans-qcqp-sol}
 	\begin{aligned}	
-		L(\mathbf{x}, \lambda) =  \sum_{i=1}^{N} w_i ||\mathbf{d}_i - \mathbf{x}||_2^2 + \lambda (
-		\mathbf{x}^\transposed \mathbf{x} - n_r^2)\\
+		\mathbf{x}^* &= \frac{1}{\lambda^* + \sum_{i=1}^{N}  w_i} \sum_{i=1}^{N} w_i \mathbf{d}_i 
 	\end{aligned}
@@
 
-then, we set the first partial derivatives to zero: 
-\begin{equation}
-	\begin{aligned}	
-		\frac{\partial L}{\partial \mathbf{x}} &= \frac{\partial }{\partial \mathbf{x}} \left( \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)^\transposed \left(\mathbf{d}_i - \mathbf{x}\right) + \lambda (
-		\mathbf{x}^\transposed \mathbf{x} - n_r^2) \right)\\
-		&= - 2\sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x}\\
-		&= \mathbf{0}\\
-	\end{aligned}
-\end{equation}
-
-By rearranging, we can obtain therefore the solution in closed form:
-\begin{equation}
-	\label{eq:trans-qcqp-lagrangian}
-	\begin{aligned}	
-		-2  &\sum_{i=1}^{N}  w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x} = \mathbf{0}\\
-		\iff \lambda \mathbf{x} &= \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)\\
-		\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \sum_{i=1}^{N}  w_i \mathbf{x}\\
-		\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \mathbf{x} \sum_{i=1}^{N}  w_i\\
-		\left( \lambda + \sum_{i=1}^{N}  w_i \right) \mathbf{x} &= \sum_{i=1}^{N} w_i \mathbf{d}_i \\
-		\mathbf{x} &= \frac{1}{\lambda + \sum_{i=1}^{N}  w_i} \sum_{i=1}^{N} w_i \mathbf{d}_i 
-	\end{aligned}
-\end{equation}
-
-Now the Karush-Kuhn-Tucker (KKT) conditions require $\lambda \geq 0$ and $\lambda (
-\mathbf{x}^\transposed \mathbf{x} - n_r^2) = 0$ in addition to satisfying the constraint, i.e.
-$|| \mathbf{x}||_2 \leq n_r$.
+where the optimal Lagrange multiplier $\lambda^*$ is chosen so that the norm of $\mathbf{x}^*$ is no greater than $n_r$ (see App. for proof).
 It follows that in practice we obtain the optimum by first computing $\mathbf{x}$ with Eq.

### 2025-05-10 sec/X_suppl.tex
Some polishing
Moving TLS relax

@@
 	\end{equation}
-	where $\mathbf{n}$ is a unit-vector that is the axis of rotation and $\theta$ is the angle of rotation.
+	where $\mathbf{n} \in \Rthree$ is a unit-vector that is the axis of rotation and $\theta$ is the angle of rotation.
 	We rewrite the objective function $\mathbf{q}^\transposed \mathbf{K} \mathbf{q}$  of \ref{eq:davenports-q-problem} using the blocks of the block-matrix $\mathbf{K}$:
@@
 \end{proof}
+
+\begin{proof}
+	We first form the Lagrangian to account for the constraint:
+	\begin{equation}
+		\begin{aligned}	
+			L(\mathbf{x}, \lambda) =  \sum_{i=1}^{N} w_i ||\mathbf{d}_i - \mathbf{x}||_2^2 + \lambda (
+			\mathbf{x}^\transposed \mathbf{x} - n_r^2)\\
+		\end{aligned}
+	\end{equation}
+
+	then, we set the first partial derivatives to zero: 
+	\begin{equation}
+		\begin{aligned}	
+			\frac{\partial L}{\partial \mathbf{x}} &= \frac{\partial }{\partial \mathbf{x}} \left( \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)^\transposed \left(\mathbf{d}_i - \mathbf{x}\right) + \lambda (
+			\mathbf{x}^\transposed \mathbf{x} - n_r^2) \right)\\
+			&= - 2\sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x}\\
+			&= \mathbf{0}\\
+		\end{aligned}
+	\end{equation}
+
+	By rearranging, we can obtain therefore the solution in closed form:
+	\begin{equation}
+		\label{eq:trans-qcqp-lagrangian}
+		\begin{aligned}	
+			-2  &\sum_{i=1}^{N}  w_i \left(\mathbf{d}_i - \mathbf{x}\right) + 2 \lambda \mathbf{x} = \mathbf{0}\\
+			\iff \lambda \mathbf{x} &= \sum_{i=1}^{N} w_i \left(\mathbf{d}_i - \mathbf{x}\right)\\
+			\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \sum_{i=1}^{N}  w_i \mathbf{x}\\
+			\lambda \mathbf{x} &= \sum_{i=1}^{N}  w_i \mathbf{d}_i  - \mathbf{x} \sum_{i=1}^{N}  w_i\\
+			\left( \lambda + \sum_{i=1}^{N}  w_i \right) \mathbf{x} &= \sum_{i=1}^{N} w_i \mathbf{d}_i \\
+			\mathbf{x} &= \frac{1}{\lambda + \sum_{i=1}^{N}  w_i} \sum_{i=1}^{N} w_i \mathbf{d}_i 
+		\end{aligned}
+	\end{equation}
+
+	Now the Karush-Kuhn-Tucker (KKT) conditions require $\lambda \geq 0$ and $\lambda (
+	\mathbf{x}^\transposed \mathbf{x} - n_r^2) = 0$ in addition to satisfying the constraint, i.e. $|| \mathbf{x}||_2 \leq n_r$.
+\end{proof}
